Rab10 regulates neuropeptide release by maintaining Ca2+ homeostasis and protein synthesis

Version 1

Abstract

Abstract

Dense core vesicles (DCVs) transport and release various neuropeptides and neurotrophins that control diverse brain functions, but the DCV secretory pathway remains poorly understood. Here, we tested a prediction emerging from invertebrate studies about the crucial role of the intracellular trafficking GTPase Rab10, by assessing DCV exocytosis at single- cell resolution upon acute Rab10 depletion in mature mouse hippocampal neurons, to circumvent potential confounding effects of Rab10’s established role in neurite outgrowth.

We observed a significant inhibition of DCV exocytosis in Rab10-depleted neurons, whereas synaptic vesicle exocytosis was unaffected. However, rather than a direct involvement in DCV trafficking, this effect was attributed to two ER-dependent processes, ER-regulated intracellular Ca2+ dynamics and protein synthesis. Gene ontology analysis of differentially expressed proteins upon Rab10 depletion identified substantial alterations in synaptic and ER/ribosomal proteins, including the Ca2+-pump SERCA2. In addition, ER morphology and dynamics were altered, ER Ca2+ levels were depleted and Ca2+ homeostasis was impaired in Rab10-depleted neurons. However, Ca2+ entry using a Ca2+ ionophore still triggered less DCV exocytosis. Instead, leucine supplementation, which enhances protein synthesis, largely rescued DCV exocytosis deficiency. We conclude that Rab10 is required for neuropeptide release by maintaining Ca2+ dynamics and regulating protein synthesis. Furthermore, DCV exocytosis appeared more dependent on (acute) protein synthesis than synaptic vesicle exocytosis.

Introduction

Dense core vesicles (DCVs) transport and release neuromodulators (neuropeptides, neurotrophic factors, and catecholamines) that play crucial roles in modulating diverse brain functions, including sleep, mood, memory, and learning (e.g., Cawley et al., 2016; Malva et al., 2012; Poo, 2001; Salio et al., 2006). Deficits in neuropeptide signaling pathways have been associated with several human disorders and diseases, including anxiety, depression, and obesity (Beck, 2000; Sah and Geracioti, 2013; Barde et al., 2022). Neuropeptides are synthesized in the endoplasmic reticulum (ER) and subsequently packaged into immature DCVs in the Golgi complex (Tooze et al., 2001). DCVs are transported along the neurites and undergo activity-dependent membrane fusion with the plasma membrane to release their content (Farina et al., 2015; Heidelberger et al., 1994; Nassal et al., 2022; Thomas et al., 1993; van de Bospoort et al., 2012). Despite the critical role of neuropeptides in brain functions, the regulatory mechanisms governing their secretion are not fully understood. We have shown previously that Rab3 is an essential regulator in the last steps of the DCV secretory pathway in mammalian neurons (Persoon et al., 2019). However, studies in invertebrates have also implicated other Rab proteins, including Rab2, Rab5, and Rab10, in the DCV secretory pathway: (Ailion et al., 2014; Azouz et al., 2014; Edwards et al., 2009; Hannemann et al., 2012; Lund et al., 2020; Sasidharan et al., 2012; Sumakovic et al., 2009).

Among these Rab proteins, Rab10 deficiency produces the strongest inhibition of neuropeptide accumulation in coelomocytes in C. elegans, suggesting deficits in DCV exocytosis (Sasidharan et al., 2012). With its subcellular localization in many vesicular organelles, such as plasmalemmal precursor vesicles (PPVs), GLUT4 transport vesicles, and recycling endosomes, Rab10 regulates various aspects of intracellular membrane trafficking, including vesicle formation, transport, and fusion (Chen et al., 2006; Larance et al., 2005; Mîinea et al., 2005; Sano et al., 2007; Taylor et al., 2015). Rab10 deficiency leads to deficits in these pathways, resulting in impaired neuronal outgrowth and disrupted retrograde axonal transport of signaling factors (Lazo and Schiavo, 2023). How these deficits relate to the strong inhibition of DCV exocytosis remains unknown.

Deletion of Rab10 expression or inhibiting its functions by overexpression of an inactive mutant (Rab10T23N) leads to abnormal ER morphology (English and Voeltz, 2013; Lv et al., 2015; Shih and Hsueh, 2016). Since the ER is a crucial organelle involved in protein synthesis, Ca2+ buffering, and lipid metabolism, alterations in its structure and function can directly or indirectly affect neuronal secretory pathways. Indeed, several studies have suggested the roles of the ER in the DCV pathway, such as the involvement of ER stress and lipid levels in DCV production in C. elegans (Laurent et al., 2018; Valadas et al., 2018), and the roles of ER Ca2+ as an internal Ca2+ source regulating somatodendritic dopamine release in mouse substantia nigra neurons (Patel et al., 2009). Additionally, Rab10 mutations, altered expression levels, or phosphorylation states are firmly associated with CNS disorders (Agola et al., 2011; Cheng et al., 2005; Ishida et al., 2016; Kiral et al., 2018). Hence, the strong inhibition of neuropeptide accumulation in coelomocytes in nematode Rab10 mutants may be a direct or indirect effect of Rab10 loss of function on DCV exocytosis and Rab10-dependent disease processes may (in part) be explained by DCV exocytosis impairment.

In the present study, we investigated the involvement of Rab10 in regulated secretion of neuropeptides in mammalian neurons. We directly assessed DCV exocytosis at single-cell resolution in mouse hippocampal neurons and confirmed that Rab10 is a crucial regulator of DCV exocytosis also in mammalian systems, while not affecting synaptic vesicle (SV) exocytosis. However, instead of having a direct role in the DCV secretory pathway, we observed that Rab10 is involved in DCV exocytosis through ER-dependent processes, especially reduced ER Ca2+ homeostasis and impaired global protein synthesis.

Supplementation with leucine known to boost protein synthesis, rescued the defects in DCV exocytosis. We therefore conclude that Rab10 plays a central role in regulating DCV exocytosis by maintaining Ca2+ homeostasis and protein synthesis.

Results

Rab10 regulates neuronal outgrowth but is dispensable for synaptogenesis and SV exocytosis

To study the role of Rab10 in regulated secretion, we utilized a knockdown strategy since complete knockout of Rab10 results in lethality at cell and organismal levels (Lv et al., 2015). We selected two specific shRNA sequences, shRNA#9, and shRNA#11, to deplete Rab10 expression, and a scrambled sequence as control. In mouse cortical neurons infected with either shRNA#9 or shRNA#11 at day in vitro 0 (DIV0), we observed a 75-95% decrease in Rab10 expression at DIV14 (Figure 1A). Previous studies have shown that Rab10 regulates neuronal outgrowth (Fukuda, 2008; Hutagalung and Novick, 2011). Consistent with these findings, we observed a significantly reduced dendrite and axon length in neurons infected with shRNA#9 at DIV0 (Figure 1B-D, 1G).

Rab10 is required for neuronal outgrowth but dispensable for SV exocytosis.

(A) Representative immunoblotting showing knockdown and rescue of Rab10 expression in cultured primary neurons infected with shRNA against Rab10 or rescue constructs (upper) and quantification of Rab10 levels (bottom).

(B) Example images of control or Rab10 KD hippocampal neurons (DIV14) stained for the dendrite marker MAP2 (blue), the synapse marker Syp1 (red), and the axonal marker SMI312 (green). Scale bar: 50 μm (upper) and 10 μm (bottom).

(C) Quantification of the dendritic length (MAP2).

(D) Quantification of the axonal length (SMI312).

(E) Quantification of Syp1 intensity per synapse per neuron

(F) Quantification of the Syp1-positive synapse density in MAP2-positive dendrites.

(G) Sholl analysis showing the mean number of dendritic branches against the distance from the soma.

(H) Example neurons infected with the SV fusion marker SypHy (upper), typical kymographs of neurites showing SypHy intensity increase during stimulation and upon NH4Cl superfusion (bottom).

(I) The average signal SypHy from active synapses, normalized from baseline to maximum fluorescence upon NH4Cl superfusion.

(J) SV exocytosis determined as the ratio of the maximum SypHy intensity during stimulation to the maximum during NH4Cl stimulation.

(K) SV endocytosis determined as the SypHy signal decay time constant τ in the 60 s after field stimulation.

All Data are plotted as mean±s.e.m. (A) N = 4, n = 4, one sample t-test. (C-G) Control: N = 3, n = 35; ShRNA#9: N = 3, n = 32. (J, K) Control: N = 3, n = 56; ShRNA#9: N = 3, n = 56. (C-F, J, K) Mann-Whitney test. **** = p<0.0001, ** = p<0.01, ns = not significant.

To test the effects of Rab10 depletion on synaptogenesis, synaptic vesicles (SVs) were quantified using the endogenous marker synaptophysin-1 (Syp1). Syp1 staining exhibited a punctate distribution at DIV14, indicating the accumulation of SVs in boutons/synapses, and no changes in the number of puncta per µm neurite or the intensity of Syp1 puncta (Figure 1E-F). These data confirm that Rab10 regulates neurite outgrowth, but we found no evidence for a role in synaptogenesis.

To test whether Rab10 depletion affects SV exocytosis, hippocampal neurons were infected with the SV exocytosis reporter Synaptophysin-pHluorin (SypHy; Figure 1H; Granseth et al., 2006). SV exocytosis was triggered by high-frequency electrical stimulation (HFS, 5s 40- Hz). The total vesicle pool was measured by briefly superfusing Tyrode’s solution containing 50 mM NH4Cl. The fraction of fused SVs, determined by the ratio of SypHy intensity upon HFS to the maximum intensity upon NH4Cl superfusion, was comparable in the two groups (Figure 1I-J). In addition, SV endocytosis, measured by the fluorescence decay of SypHy after HFS, was unaffected by Rab10 depletion (Figure 1K). Therefore, we conclude that Rab10 is dispensable for SV exocytosis.

Rab10 is a major regulator of DCV exocytosis

To investigate the role of Rab10 in neuropeptide release, we depleted Rab10 levels at DIV0, expressed the DCV exocytosis reporter NPY-pHluorin at DIV9-10, and performed live-cell imaging at DIV14. Our previous studies have demonstrated that NPY-pHluorin almost exclusively localizes to DCVs as indicated by its strong co-localization with endogenous markers of DCVs, such as BDNF and the chromogranins CHGA and CHGB (Arora et al., 2017; Dominguez et al., 2018; Persoon et al., 2019). To achieve single-vesicle resolution analysis of DCV exocytosis, we used single cultured hippocampal neurons on glial micro- islands. Neurons were stimulated by 16 trains of 50 action potentials (APs) at 50 Hz, a protocol known to trigger robust DCV exocytosis (Balkowiec and Katz, 2002; Emperador-Melero et al., 2018; Gärtner and Staiger, 2002; Hartmann et al., 2001; Moro et al., 2021; Persoon et al., 2019). Fusion events were detected as a rapid appearance of fluorescent puncta (Figure 2—figure supplement 1A-B). DCV exocytosis in Rab10 KD neurons was significantly reduced by 60% compared to control neurons (Figure 2—figure supplement 1C- D). Furthermore, the total number of DCVs was reduced by 30% in Rab10 KD neurons (Figure 2—figure supplement 1E). The fusion fraction, which represents the number of DCV fusion events relative to the remaining DCV pool after stimulation, was also significantly decreased by 50% in Rab10 KD neurons (Figure 2—figure supplement 1F). Overexpression of wild-type, knockdown-resistant Rab10 rescued DCV exocytosis deficits in Rab10 KD neurons (Figure 2—figure supplement 1C-F).

To overcome the potential confounding effects of impaired neurite outgrowth (Figure 1) and reduced total DCV pool (Figure 2—figure supplement 1) upon Rab10 depletion starting at DIV0, we adopted a more acute approach to interfere with Rab10 function, and late enough not to affect neuronal morphology and the total DCV pool. Neurons were transfected with shRNA against Rab10 at DIV7, fixed at DIV14, and stained with markers for dendrites (MAP2), axons (SMI312), and SVs (Syp1). Rab10 expression was reduced by 70% after 7 days of infection (Appendix 1—figure 3). No significant alterations in total dendrite length, axon length (Figure 2A-C, 2E), or synapse density (Figure 2D) were observed in Rab10 KD neurons under these conditions. Therefore, to eliminate confounding effects on morphological parameters, we reevaluated DCV exocytosis and all further experiments in neurons infected with shRNA at DIV7. DCV exocytosis in Rab10 KD neurons remained significantly lower by 50% compared to control neurons (Figure 2F-I). The fusion fraction was also significantly reduced by 65% in Rab10 KD neurons (Figure 2K). Overexpression of wild-type Rab10 rescued DCV exocytosis deficits. No significant differences in the total number of DCVs (Figure 2J), DCV transport (Figure 2—figure supplement 2A-2B, 2E), or cargo loading (Figure 2—figure supplement 2F-K) were observed. Thus, these data indicate that Rab10 depletion specifically inhibits activity-dependent neuropeptide release in hippocampal neurons, without effects on DCV biogenesis, cargo loading, and transport and independent of Rab10’s established role in neuronal outgrowth.

Rab10 is a major regulator of DCV exocytosis.

(A) Example images of control and Rab10 KD hippocampal neurons (DIV14) stained for MAP2 (blue), Syp1 (red) and SMI312 (green). Scale bar: 30 μm.

(B) Quantification of the dendritic length (MAP2).

(C) Quantification of the axonal length (SMI312).

(D) Quantification of the Syp1-positive synapse density in MAP2-positive dendrites.

(E) Sholl analysis showing the mean number of dendritic branches against the distance from the soma.

(F) Schematic representation of DCV fusion assay. DCVs are labeled with NPY-pHluorin, and neurons are stimulated with one train of 16 bursts of 50 APs at 50 Hz (light blue bars).

(G) Representative neurons during electrical stimulation superimposed with NPY-pHluorin fusion events (green dots). Scale bar: 5 μm.

(H) Cumulative plot of DCV fusion events per cell. Light blue bars represent the stimulation trains.

(I) Summary graph of DCV fusion events per cell.

(J) The total number of DCVs (total pool) of neurons analyzed in Figure 2H, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

(K) Fraction of NPY-pHluorin-labeled DCV fusing during stimulation.

All Data are plotted as mean±s.e.m. (B-D) Control: N = 3, n = 31; ShRNA#9: N = 3, N = 28; ShRNA#11: N = 3, n = 31. (I-K) Control: N = 4, n = 36; shRNA#9: N = 4, N = 37; shRNA#11: N =4, n = 30; Rescue: N = 4, n = 34. (B, C, I, K) Kruskal-Wallis test. (D, J) one-way analysis of variance (ANOVA). **** = p<0.0001, *** = p<0.001, ** = p<0.01, * = p <0.05, ns = not significant.

Proteins involved in synaptic transmission and translation are severely dysregulated upon Rab10 depletion

To comprehensively investigate Rab10 function in mature neurons, mass spectrometry proteomics was performed on Rab10 KD and control neurons at DIV14. A total of approximately 5400 unique proteins were identified and quantified in two biological replicates. The complete list of proteins quantified in this study is presented in Appendix 1- figure 3. Only differentially expressed proteins detected with high confidence characterized by a log2(fold change) > 0.56, and q-value < 0.01, were included in the subsequent analysis. Among the dysregulated proteins, 71% were upregulated, while 29% were downregulated, resulting in a significant dysregulation of 19% of the total protein pool in Rab10 KD neurons. These data indicate that Rab10 depletion leads to major neuronal proteome dysregulation within 7 days after initiating knockdown (Figure 3A).

Depletion of Rab10 leads to dysregulation of proteins enriched in presynaptic transmission and cytosolic translation.

(A) Volcano plots showing significantly dysregulated proteins in Rab10 depleted neurons.

(B) GO enrichment analysis of functional pathways of the significant hits with ClueGO. Shown are the Bonferroni corrected p-values.

(C) GO enrichment analysis of subcellular localization of the significant hits with ClueGO. Shown are the Bonferroni corrected p-values.

(D) Sunburst plot showing the annotation in synaptic function of the altered proteins in Rab10 depleted neurons.

(E) Sunburst plot showing the annotation in synaptic location of the altered proteins in Rab10 depleted neurons.

(F) Log2 fold changes of synaptic proteins within SynGO terms. Downregulated proteins are shown in blue and upregulated proteins are shown in black.

(G) Examples of proteins that are significantly affected by Rab10 depletion grouped by their subcellular localization. Heat maps represent the degree of up- or downregulation.|

(H) Selective MS data analysis of ER-related proteins in Rab10 KD neurons. Bars show the fold change of the indicated peptides compared to the control.

To gain insights into the functional consequences of Rab10 depletion, we performed Gene Ontology (GO) analysis using ClueGO (Bindea et al., 2009). This analysis revealed that biological processes related to chemical synaptic transmission were notably affected by Rab10 depletion (Figure 3B). In addition, several biological processes related to protein synthesis, such as cytoplasmic translation and ribosomal large subunit biogenesis, were among the top 5 terms with the lowest p-values (Figure 3B). Subcellular localization analysis of these dysregulated proteins upon Rab10 depletion showed that cytosolic ribosomal proteins were the most significantly affected, followed by dendritic proteins (Figure 3C).

Further characterization of the dysregulated synaptic proteins was performed using SynGO (Koopmans et al., 2019). Among the 391 significantly dysregulated proteins annotated in SynGO, 205 were classified as presynaptic proteins and 237 as postsynaptic proteins. GO enrichment analysis revealed that biological processes in metabolism were most dysregulated upon Rab10 depletion. Among them, both pre and postsynaptic translation were significantly dysregulated (Figure 3D). Consistent with the ClueGO analysis, SynGO highlighted a significant dysregulation of presynaptic (34 of 391) and postsynaptic ribosomal proteins (44 of 391), supporting the involvement of Rab10 in the regulation of neuronal protein synthesis (Figure 3E). Interestingly, all dysregulated ribosomal proteins were upregulated upon Rab10 depletion, which contrasts with the mostly downregulated expression observed in the other classes of proteins (such as synaptic and cytoskeletal proteins) (Figure 3F-G). Taken together, GO analysis with both ClueGO and SynGO indicates a dysfunction of protein translation in Rab10 KD neurons.

Loss of Rab10 expression has been associated with altered ER morphology in mouse embryonic cells (Lv et al., 2015), which may explain the selective upregulation of proteins involved in ribosome function in our proteomics data. Indeed, ER proteins were dysregulated substantially upon Rab10 depletion. Specifically, several rough ER proteins showed differential regulation, with SEC61A1 being upregulated, SEC61G being downregulated, and CLIMP remaining unchanged (Figure 3H). Most tubular ER proteins, such as RTN3/4 and VAPB, were robustly decreased. Interestingly, one of the ER membrane Ca2+ channels, SERCA2, showed a 50% reduction upon Rab10 depletion (Figure 3H).

Taken together, these analyses reveal that Rab10 depletion leads to major changes in protein expression, especially synaptic and ribosomal proteins, the latter all upregulated in Rab10- depleted neurons which suggests that protein synthesis is dysregulated, potentially due to altered ER function.

Rab10 regulates ER morphology and ribosomal protein levels

Given the substantial dysregulation of synapse and ribosome/ER proteins, we investigated synapses and ER further using electron microscopy (Figure 4). These analyses revealed an apparently normal synapse morphology in Rab10 depleted neurons with many SVs clustered at the active zone, while DCVs were sparsely distributed along neurites and near the active zone (Figure 4A). The length of the active zone (AZ) and postsynaptic density (PSD) were both decreased by 10% upon Rab10 depletion (Figure 4C-D). However, other parameters of synaptic ultrastructure, such as the diameter of SVs or DCVs, and the number of SVs per synapse, were unchanged in Rab10 KD neurons (Figure 4E-H). Hence, despite substantial dysregulation of synaptic proteins, overall synapse morphology was hardly affected.

Rab10 regulates synapse size and ER morphology.

(A) Representative EM pictures showing the ultrastructure of synapses. Scale bar 100 nm.

(B) Representative EM pictures showing the ultrastructure of soma. Rough ER (rER) is indicated by red dotted lines. M: mitochondrion, G: Golgi. Scale bar: 100 nm.

(C) Quantification of the length of active zone and postsynaptic density (PSD).

(D) Quantification of the length of postsynaptic density (PSD).

(E) Quantification of SV number per synapse and SV diameter.

(F) Quantification of SV diameter.

(G) Quantification of DCV diameter.

(H) Frequency distribution of DCVs by diameter.

(I) Quantification of the diameter of rough ER (rER).

Data are plotted with Superplot (Figure. 4C-G, I), where averages from three independent cultures are shown as large circles and single observations are shown as dots. Horizontal lines represent the means of the averages from three weeks. Data from different cultures are grouped with different colors. (C- D) Control: N = 3, n = 184; shRNA#9: N = 3, n = 187. (E) Control: N = 3, n = 189; shRNA#9: N = 3, n = 188. (F) Control: N = 3, n = 1770; shRNA#9: N = 3, n = 1803. (G) Control: N = 3, n = 137; shRNA#9: N = 3, n = 122. (I) Control: N = 3, n = 63; shRNA#9: N = 3, n = 64. (C-G, I) linear mixed model analysis. *** = p<0.001, * = p <0.05, ns = not significant.

Tubular ER was also observed in some presynaptic sections, consistent with previous studies (Deng et al., 2021; Droz et al., 1975; Wu et al., 2017). The percentage of synaptic sections containing tubular ER was decreased by 24% (37% in Control versus 28% in Rab10 KD).

Rough ER (rER) was also studied in the somata of Rab10 depleted and control neurons (Figure 4I). The diameter of rER tubes was reduced by 15% in Rab10 KD neurons. Hence, the substantial dysregulation of ribosomal and ER proteins in Rab10-depleted neurons was accompanied by changes in the abundance of synaptic ER and small changes in ER morphology in the soma.

To study these effects on ER abundance and morphology further, we performed immunofluorescence staining for two endogenous ER markers, KDEL and RTN4. The average fluorescence intensity of RTN4 and KDEL staining was significantly decreased by 35% and 25% respectively in Rab10 KD neurons (Figure 4—figure supplement 1A-C). The relative distributions of RTN4 and KDEL in neurites, as calculated by the intensity ratio of these two proteins in neurites over their somatic intensity, were reduced by 25% and 13%, respectively (Figure 4—figure supplement 1D-E). In conclusion, the ultrastructural changes in ER abundance and morphology upon Rab10 depletion were accompanied by altered distribution of axonal ER, without affecting the ultrastructure of SVs and DCVs.

Finally, given the dynamic nature of ER tubular networks and the involvement of Rab10 in ER tubule extension in COS- 7 cells (English and Voeltz, 2013), we tested ER dynamics in Rab10 KD and control neurons expressing the luminal ER marker mCherry-ER3 using live- cell imaging at DIV8, and performed FRAP experiments at DIV14. The recovery of mCherry-ER3 intensity after photobleaching was significantly slower in Rab10 KD neurons with only 50% recovery within three minutes, compared to 80% recovery in control neurons (Figure 4—figure supplement 2C). Collectively, these data indicate that Rab10 depletion leads to reduced levels of ER-resident proteins altered ER abundance and morphology, and impaired ER dynamics.

Rab10 regulates SERCA2 levels and ER Ca2+ homeostasis

The ER is the largest internal Ca2+ source in neurons and plays a crucial role in maintaining neuronal Ca2+ homeostasis (Karagas and Venkatachalam, 2019). The maintenance of ER Ca2+ primarily relies on the Sarco Endoplasmic Reticulum Calcium ATPase (SERCA), a Ca2+ pump. Among the three major paralogs of SERCA, SERCA2 is particularly enriched in neurons (Britzolaki et al., 2018; Periasamy and Kalyanasundaram, 2007; Xu and Van Remmen, 2021). Consistent with the proteomic analysis which revealed a reduced SERCA2 expression upon Rab10 depletion (Figure 3), immunoblotting confirmed the reduction of SERCA2 levels, showing a 50% reduction in Rab10 KD neurons (Figure 5A-5B). Therefore, ER alternations in Rab10 KD neurons may disrupt Ca2+ homeostasis, which is essential for DCV exocytosis.

Reduced SERCA2 levels and impaired ER Ca2+ homeostasis in Rab10 KD neurons.

(A) Typical immunoblot showing reduced SERCA2 levels in Rab10 KD hippocampal neurons.

(B) Quantification of protein levels in Rab10 KD neurons normalized to control.

(C) Quantification of somatic ER Ca2+ concentration.

(D) Quantification of dendritic ER Ca2+ concentration.

(E) Representative image of a neuron infected with ER-GCaMP6-150 displayed with a pseudo line. Scale bar: 3 μm.

(F) Typical kymographs of the somatic intensity of ER-GCaMP6-150 showing the intensity decrease upon caffeine superfusion (red line), and the recovery in intensity after caffeine washout. Scale bar: 10 s.

(G) Average normalized ER-GCaMP6-150 fluorescence recovery after caffeine treatment.

(H) Normalized ER-GCaMP6-150 fluorescence recovery after caffeine treatment at T = 190 s. All Data are plotted as mean±s.e.m. (B) Control: N = 4, n = 4; Rab10 KD: N = 4, n = 4; (C-D) Control: N = 3, n = 17; Rab10 KD: N = 3; n = 17; Rescue: N = 3, n = 17 (F) Control: N = 3, n = 23; Rab10 KD: N = 3; n = 24; Rescue: N = 3, n = 24. (B) One sample t-test. (C, D, F) Kruskal-Wallis test.

**** = p<0.0001, *** = p<0.001, ** = p<0.01, ns = not significant.

To test this, we next measured the Ca2+ concentration in ER ([Ca2+]ER) using the ER Ca2+ indicator ER-GCaMP6 (de Juan-Sanz et al., 2017). We observed a reduction in [Ca2+]ER in soma from 130 μM in control neurons to about 70 μM in Rab10-depleted neurons (Figure 5C). [Ca2+]ER in neuritis was also reduced by 15% in Rab10 KD neurons (Figure 5D). The reduction of [Ca2+]ER was rescued by the expression of a knockdown-resistant Rab10 construct. To validate this observation, we assessed ER Ca2+ homeostasis indirectly by measuring the effect of caffeine on cytosolic Ca2+ concentration. As expected, in the absence of extracellular Ca2+, caffeine application (1 μM) triggered an increase in cytosolic Ca2+ due to Ca2+ release from the ER in both wild-type and Rab10 depleted neurons. However, the peak and the area of the caffeine-induced Ca2+ response curves were both reduced by 30% in Rab10 KD neurons (Figure 5—figure supplement 1).

Furthermore, we examined ER Ca2+ dynamics following a 10-minute caffeine treatment. Caffeine activates the ryanodine receptor (RyR), leading to the depletion of ER Ca2+ (Endo, 1975; Fujimoto et al., 1980). As expected, perfusion with caffeine-induced ER Ca2+ depletion, followed by recovery towards pre-stimulation levels (Figure 5F-G). In wild- type neurons, Ca2+ levels were recovered by 90% at T=190 s. However, the refilling of ER Ca2+ was significantly delayed in Rab10 KD neurons, and Ca2+ levels were only recovered by 50% at T=190 s (Figure 5G-H).

Finally, to investigate the consequence of the ER Ca2+ depletion on neuronal Ca2+ homeostasis in Rab10 KD neurons, we measured cytosolic Ca2+ responses triggered by action potentials (AP) using the Ca2+ indicator Fluo5F (Figure 6A-C) and the genetically encoded Synaptophysin-GCaMP6 (Figure 6D-F). The AP-induced Ca2+ responses in the soma, as measured by Fluo5F, were reduced by 40% upon Rab10 depletion (Figure 6C). Similarly, the AP-induced Ca2+ responses in presynaptic nerve terminals, measured by Synaptophysin- GCaMP6, were also reduced (20%, Figure 6F), although this effect was smaller than the effects of Rab10 KD on ER Ca2+ levels and Caffeine-induced ER Ca2+ depletion (Figure 5). Taken together, these data suggest that Rab10 knockdown leads to ER Ca2+ depletion and impairs neuronal Ca2+ homeostasis, which may be attributed to the reduced levels of SERCA2 level and slower ER Ca2+ refilling and may contribute to the impaired DCV exocytosis.

Impaired neuronal Ca2+ influx triggered by electrical stimulation.

(A) Representative time-lapse of cytosolic Fluo-5 AM upon electrical stimulation (16 APs, 50Hz) in somas of hippocampal neurons. Scale bar: 10 μm.

(B) Average normalized response of somatic Fluo-5 AM fluorescence upon stimulation (16 APs, 50Hz) in hippocampal neurons.

(C) Quantification of the area under the curve (AUC) of the Fluo-5 AM fluorescence traces.

(D) Typical neurons infected with Synaptophysin-GCaMP6 (upper), typical kymograph of a neurite (bottom) showing Synaptophysin-GCaMP6 intensity increase upon electrical stimulation (16 APs, 50Hz, blue bars). Scale bar: 5μm.

(E) Average normalized response of Synaptophysin-GCaMP6 fluorescence intensity at presynaptic boutons upon stimulation (16 APs, 50Hz) in hippocampal neurons.

(F) Quantification of the area under the curve (AUC) of the Synaptophysin-GCaMP6 fluorescence traces in control and Rab10 KD neurons.

All Data are plotted as mean±s.e.m. (C) Control: N = 4, n = 24; Rab10 KD: N = 4, n = 30; Rescue: N = 4, n = 27. (F) Control: N = 3, n = 33; Rab10 KD: N = 3; n = 27. (C) Kruskal-Wallis test. (F) Unpaired t-test. ** = p<0.01, * = p <0.05, ns = not significant.

Rab10 depletion impairs ionomycin-induced DCV exocytosis

To determine whether the impaired Ca2+ signaling explains the DCV exocytosis deficiency in Rab10 KD neurons, we stimulated DCV exocytosis using the Ca2+ ionophore ionomycin.

This approach bypasses cellular Ca2+ homeostasis and artificially increases the intracellular Ca2+ concentration enough to trigger DCV exocytosis (Persoon et al., 2019). Unexpectedly, DCV exocytosis was still reduced in Rab10 KD neurons (Figure 7A-B). Although the reduction was substantial, 45%, this impairment was not as substantial as observed for action potential-induced DCV exocytosis (65%, Figure 2K). The total number of DCV was nearly identical between control and Rab10 KD neurons (Figure 7D). Thus, although a minor fraction of the DCV exocytosis deficits may be explained by impaired Ca2+ signaling (difference between 45% and 65%), other deficits explain most of the DCV exocytosis deficiency in Rab10 KD neurons.

Impaired DCV fusion induced by ionomycin in Rab10 KD neurons.

(A) Representative neurons during electrical stimulation superimposed with NPY- pHluorin fusion events (green dots). Scale bar: 10 μm.

(B) Cumulative plot of DCV fusion events per cell.

(C) Fraction of NPY-pHluorin-labeled DCVs fusing during stimulation.

(D) The total number of DCVs (total pool) of neurons analyzed in Figure 7B, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

All Data are plotted as mean±s.e.m. (C, D) Control: N = 3, n = 20; Rab10 KD: N = 3, n = 21. (C, D) Mann-Whitney test. * = p <0.05, ns = not significant.

Rab10 regulates neuronal protein synthesis

Since significant dysregulation of ER markers and ribosomal proteins was observed upon Rab10 depletion, we investigated the effects of Rab10 on protein synthesis using SUnSET to detect nascent peptides formed during puromycin pulse labeling (Schmidt et al., 2009).

SUnSET analysis revealed that global protein synthesis was reduced by 30% upon Rab10 depletion (Figure 8A-B).

Leucine supplementation ameliorates the deficits in protein synthesis and neuropeptide release in Rab10 KD neurons.

(A) Representative western blot showing puromycinilated proteins as a measure for de novo protein synthesis in each condition.

(B) Quantification of puromycin intensity in each condition.

(C) Representation of the DCV fusion assay. Leucine (5 μM) was added to the culture media and incubated for 72 h before DCV fusion assay. DMSO (1‰) was used as a control.

(D) Cumulative plot of DCV fusion events per cell.

(E) Fraction of NPY-pHluorin-labeled DCVs fusing during stimulation.

(F) The total number of DCVs (total pool) of neurons analyzed in Figure 8D, E, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

All Data are plotted as mean±s.e.m. (B) All: N = 3, n = 3 (E, F) Control: N = 3, n = 47; Control + leu: N = 3, n = 45; Rab10 KD: N = 3; n = 61; Rab10 + leu: N = 3, n = 54. Rab10 KD + Rab10: N = 3, n = 24. (B) One sample t-test. (E, F) Kruskal-Wallis test. ** = p<0.01, * = p <0.05, ns = not significant.

Protein synthesis impairments may be rescued by supplementation with leucine, a branched- chain amino acid, that promotes protein synthesis by activating the mTOR pathway (Ananieva et al., 2016). To test this in Rab10 KD neurons, additional L-leucine was added to the culture medium to increase the concentration to 5 mM for three days. Indeed, 5mM leucine supplementation significantly restored global protein synthesis deficits caused by Rab10 depletion (Figure 8A-B).

Finally, a similar impairment in global protein synthesis was observed when a loss-of- function mutant of Rab10 (Rab10T23N) was overexpressed in wild-type neurons (Figure 8A- B). Thus, Rab10 depletion or Rab10T23N expression reduces global protein synthesis in neurons, probably by dysregulation of ER and ribosomal function.

Leucine supplementation restores normal DCV exocytosis

We hypothesized that protein synthesis deficits in Rab10-depleted neurons explain most of the impaired DCV exocytosis (in addition to a minor fraction explained by disturbed Ca2+- homeostasis, see above) and tested whether leucine supplementation could rescue the DCV exocytosis deficits in Rab10 KD neurons. Rab10 depleted neurons expressing NPY-pHluorin were treated with 5mM leucine three days before live-cell imaging, or with dimethyl sulfoxide (DMSO) as a control. Leucine supplementation restored DCV exocytosis by 80% caused by Rab10 depletion but did not alter DCV exocytosis in control neurons (Figure 8C- F). These results suggest that impaired protein synthesis is a major factor contributing to DCV exocytosis deficits in Rab10-depleted neurons.

Discussion

In this study, we investigated the function of Rab10 in neuropeptide release in mature mouse hippocampal neurons. We found that DCV exocytosis triggered by action potential trains was reduced by 65% upon Rab10 depletion, whereas SV exocytosis was unaffected. In addition, we observed a depleted ER Ca2+ pool and an impaired action potential-induced Ca2+ response in Rab10 KD neurons. However, DCV exocytosis triggered by Ca2+-ionophore ionomycin, a triggering method independent of Ca2+ channels and internal Ca2+-stores, was also impaired, albeit to a lesser extent. Furthermore, ribosomal proteins were massively dysregulated, and protein synthesis was impeded upon Rab10 depletion. Finally, the DCV exocytosis deficit in Rab10 KD neurons was largely rescued by leucine supplementation. We conclude that the strong inhibition of DCV exocytosis upon Rab10 depletion is mostly due to protein synthesis deficiency and to a lesser extent by dysregulation of Ca2+ channels and internal Ca2+-stores.

Rab10 regulates neurite outgrowth but not membrane homeostasis in mature neurons

Rab10 is highly enriched in neurons and plays crucial roles in neuronal development (Taylor et al., 2015; Wang et al., 2011; Xu et al., 2014). Consistent with these previous findings, we observed a reduction of axonal and dendritic outgrowth in Rab10-depleted neurons. Also consistent with previous findings in invertebrates (Sasidharan et al., 2012), we observed that the endogenous levels of SV markers are unchanged, indicating that SV biogenesis was unaffected. However, morphological characterization of neurons infected with shRNAs after the first week in culture did not identify changes in the total length of dendrites or axons, indicating that Rab10 is selectively required for outgrowth, not for subsequent membrane homeostasis or cell viability in mature neurons. Therefore, the strong deficit in neuropeptide release in DIV7-infected neurons is unlikely to be confounded by Rab10-dependent aspects of neuronal development.

Rab10 is crucial for DCV exocytosis but not for SV exocytosis

In C. elegans, neuropeptide release was abolished in Rab10 deletion mutants (Sasidharan et al., 2012). We observed a 60% reduction in neuropeptide release in Rab10-depleted mature mouse hippocampal neurons. The difference in effect size could be explained by the incomplete Rab10 depletion with shRNA silencing or by redundant pathways in mammals, e.g. Rab10 and Rab8 are closely related paralogs and share many common effectors (Homma et al., 2020).

Unlike Rab3, which travels together DCVs and exhibits a reduction of over 90% in DCV exocytosis in its quadruple knock-out neurons (Persoon et al., 2019), Rab10 does not typically travel with DCVs (Figure 2—figure supplement 3). In addition, no changes in DCV size, puncta intensity, puncta distribution, travel velocity, or -distance were detected in Rab10 KD neurons. We conclude that DCV exocytosis deficiency in Rab10-depleted neurons is not caused by alterations in DCV biogenesis or transport and that Rab10 is not required for DCV trafficking.

Although Rab10 is found in subcellular fractions enriched in synaptic vesicles (Takamori et al., 2006; Taoufiq et al., 2020), it is dispensable for SV exocytosis. Evoked postsynaptic currents (EPSCs) were unaffected in Rab10 mutants in C. elegans (Sasidharan et al., 2012). In line with this, SV exocytosis was unaffected in Rab10-depleted hippocampal neurons.

These results indicate that Rab10 is selectively required for DCV exocytosis, not for SV exocytosis. Strikingly, many synaptic proteins, including many involved in SV exocytosis, are among the most severely dysregulated proteins upon Rab10 depletion. SV exocytosis may be more resilient to acute protein changes (Sinha et al., 2011) than DCVs. In addition, vesicle secretion properties are different for DCVs and SVs. Unlike SVs, which are secreted upon a single electrical stimulation, DCVs need prolonged or more intense stimulation for the induction of fusion. Thus, the regulatory effects of Rab10 on DCV exocytosis might be amplified under prolonged stimulation.

Ca2+ homeostasis deficits contribute to DCV exocytosis deficits in Rab10 KD neurons

Rab10 KD neurons showed depleted ER Ca2+ and impaired cytosolic Ca2+ responses. These effects may contribute to the observed DCV exocytosis deficits. Ca2+ released from the ER promotes DCV mobility and potentiates neuropeptide release via activating the CaMKII pathway in Drosophila (Shakiryanova et al., 2007). However, this might not be the case in mouse neurons since most axonal ER takes Ca2+ up from cytosol, instead of releasing it (de Juan-Sanz et al., 2017). Secondly, triggering Ca2+ release from the ER did not alter DCV transport and fusion (unpublished data from our lab). Finally, a previous study from our lab has shown that CaMKII deficiency does not alter DCV exocytosis (Moro et al., 2020).

Hence, dysregulation of ER Ca2+ dynamics may not directly explain the observed DCV exocytosis deficits in Rab10-depleted neurons. However, ER Ca2+ also regulates Ca2+ influx by modulating L-type voltage-gated Ca2+ channels at the plasma membrane via a STIM1- based feedback loop (de Juan-Sanz et al., 2017). Given the importance of Ca2+ influx for DCV exocytosis, dysregulation of ER Ca2+ dynamics may indirectly explain DCV exocytosis deficiency upon Rab10 depletion. However, Ionomycin-triggered DCV exocytosis, which bypasses voltage-gated Ca2+ channels, was still reduced, albeit to a lesser extent compared to action potential trains. This difference in effect size is consistent with a limited contribution for dysregulation of Ca2+ dynamics and voltage-gated Ca2+ channels to explain the impaired DCV exocytosis in Rab10 KD neurons, while the majority of this phenotype is explained by deficits downstream of protein synthesis.

A role of Rab10 in protein synthesis largely explains DCV exocytosis deficiency in Rab10 KD neurons

In line with the previous study (Lv et al., 2015), which indicated alteration in ER morphology in Rab10 KO embryonic cells. We observed impaired ER morphology and upregulated ribosomal proteins upon Rab10 depletion. The upregulation of ribosomal proteins might be a compensatory response to altered ER structure as mutation of other ER-shaping proteins, such as VCP and ALT1, causes similar ribosomal abnormalities as Rab10 depletion (Shih and Hsueh, 2016). Protein homeostasis is vital for neuronal function and deficiency in protein translation is related to several CNS disorders (Cajigas et al., 2010; Holt et al., 2019; Koga et al., 2011; Laguesse and Ron, 2020).

Dysregulation of ribosomes and ER is probably sufficient to explain the impaired protein translation in Rab10 KD neurons. Depleted ER Ca2+ may induce ER stress (Arruda and Hotamisligil, 2015; Fu et al., 2011) which may also contribute to protein translation inhibition, but ATF4 levels are unaffected in Rab10 KD neurons, suggesting that ER stress is at best limited (Figure 4—figure supplement 3A-B). In addition, Rab10 regulates the retrograde transport of TrkB signaling endosomes (Lazo and Schiavo, 2023), which may activate the CREB/mTOR pathway and promote protein synthesis (Moya-Alvarado et al., 2023). Hence, impaired TrkB transport may also contribute to impaired protein translation in addition to dysregulated ribosomes in Rab10 KD neurons. We found that restoring protein synthesis with leucine efficiently increased protein synthesis and largely restored DCV exocytosis deficiency in Rab10-depleted neurons. We conclude that dysregulation of protein synthesis results in DCV exocytosis deficits in these neurons.

In conclusion, our data demonstrate the importance of Rab10 in neuropeptide release and ER homeostasis. We observed altered ER morphology, reduced ER Ca2+ concentration, impaired protein synthesis, and impaired neuropeptide release in Rab10-depleted neurons. These observations shed light on the pathogenesis of Rab10-related disease. In addition, we have shown leucine can largely rescue deficiency in protein synthesis and neuropeptide release in Rab10-depleted neurons, providing a potential treatment for disorders associated with neuropeptide abnormalities such as depression and anxiety.

Contributions

Conceptualization: JD, RT, MV

Methodology: JD, MC, RD, JW, KWL, ABS, RT, MV

Investigation: JD, RT, MV

Visualization: JD, MC, KWL, JW, RT, MV Funding acquisition: RT, MV

Supervision: RT, MV

Writing – original draft: JD, RT, MV

eLife Assessment

Dong et al. investigate the role of the small Ras-like GTPase Rab10 in the exocytosis of DCVs in mouse hippocampal neurons, showing that Rab10 depletion hinders DCV exocytosis independently of its effects on neurite outgrowth. Their findings are convincing and provide evidence that Rab10 depletion leads to altered ER morphology, impaired ER-based calcium buffering, and decreased ribosomal protein expression, which collectively contributes to defective DCV secretion. The study comes to the important conclusion that Rab10 is critical for DCV release by ensuring ER calcium homeostasis.

Reviewer #1 (Public Review):

Summary:

Dong et al here have studied the impact of the small Ras-like GTPase Rab10 on the exocytosis of dense core vesicles (DVC), which are important mediators of neuropeptide signaling in the brain. They use optical imaging to show that lentiviral depletion of Rab10 in mouse hippocampal neurons in culture independent of the established defects in neurite outgrowth hamper DCV exocytosis. They further demonstrate that such defects are paralleled by changes in ER morphology and defective ER-based calcium buffering as well as reduced ribosomal protein expression in Rab10-depleted neurons. Re-expression of Rab10 or supplementation of exogenous L-leucine to restore defective neuronal protein synthesis rescues impaired DCV secretion. Based on these results they propose that Rab10 regulates DCV release by maintaining ER calcium homeostasis and neuronal protein synthesis.

Strengths:

This work provides interesting and potentially important new insights into the connection between ER function and the regulated secretion of neuropeptides via DCVs. The authors combine advanced optical imaging with light and electron microscopy, biochemistry, and proteomics approaches to thoroughly assess the effects of Rab10 knockdown at the cellular level in primary neurons. The proteomic dataset provided may be valuable in facilitating future studies regarding Rab10 function. This work will thus be of interest to neuroscientists and cell biologists.

Weaknesses:

While the main conclusions of this study are comparably well supported by the data, I see three major weaknesses:

(1) For some of the data the statistical basis for analysis remains unclear. I.e. is the statistical assessment based on N= number of experiments or n = number of synapses, images, fields of view etc.? As the latter cannot be considered independent biological replicates, they should not form the basis of statistical testing.

(2) As it stands the paper reports on three partially independent phenotypic observations, the causal interrelationship of which remains unclear. Based on prior studies (e.g. Mercan et al 2013 Mol Cell Biol; Graves et al JBC 1997) it is conceivable that defective ER-based calcium signaling and the observed reduction in protein synthesis are causally related. For example, ER calcium release is known to promote pS6K1 phosphorylation, a major upstream regulator of protein synthesis and ribosome biogenesis. Conversely, L-leucine supplementation is known to trigger calcium release from ER stores via IP3Rs. Given the reported impact of Rab10 on axonal transport of autophagosomes and, possibly, lysosomes via JIP3/4 or other mediators (see e.g. Cason and Holzbaur JCB 2023) and the fact that mTORC1, the alleged target of leucine supplementation, is located on lysosomes, which in turn form membrane contacts with the ER, it seems worth analyzing whether the various phenotypes observed are linked at the level of mTORC1 signaling.

(3) The claimed lack of effect of Rab10 depletion on SV exocytosis is solely based on very strong train stimulation with 200 Aps, a condition not very well suited to analyze defects in SV fusion. The conclusion that Rab10 loss does not impact SV fusion thus seems premature.

Reviewer #2 (Public Review):

Summary:

In this paper, the authors assess the function of Rab10 in dense core vesicle (DCV) exocytosis using RNAi and cultured neurons. The author provides evidence that their knockdown (KD) is effective and provides evidence that DCV is compromised. They also perform proteomic analysis to identify potential pathways that are affected upon KD of Rab10 that may be involved in DCV release. Upon focusing on ER morphology and protein synthesis, the authors conclude that defects in protein synthesis and ER Ca2+ homeostasis contributes to the DVC release defect upon Rab10 KD. The authors claim that Rab10 is not involved in synaptic vesicle (SV) release and membrane homeostasis in mature neurons.

Strengths:

The data related to Rab10's role in DCV release seems to be strong and carried out with rigor. While the paper lacks in vivo evidence that this gene is indeed involved in DCV in a living mammalian organism, I feel the cellular studies have value. The identification of ER defect in Rab10 manipulation is not truly novel but it is a good conformation of studies performed in other systems. The finding that DCV release defect and protein synthesis defect seen upon Rab10 KD can be significantly suppressed by Leucine supplementation is also a strength of this work.

Weaknesses:

The data showing Rab10 is NOT involved in SV exocytosis seems a bit weak to me. Since the proteomic analysis revealed so many proteins that are involved in SV exo/encodytosis to be affected upon Rab10, it is a bit strange that they didn't see an obvious defect. Perhaps this could have been because of the protocol that the authors used to trigger SV release (I am not an E-phys expert but perhaps this could have been a 'sledge-hammer' manipulation that may mask any subtle defects)? Perhaps the authors can claim that DCV is more sensitive to Rab10 KD than SV, but I am not sure whether the authors should make a strong claim about Rab10 not being important for SV exocytosis.

Also, the authors mention "Rab10 does not regulate membrane homeostasis in mature neurons" but I feel this is an overstatement. Since the authors only performed KD experiments, not knock-out (KO) experiments, I believe they should not make any conclusion about it not being required, especially since there is some level of Rab10 present in their cells. If they want to make these claims, I believe the authors will need to perform conditional KO experiments, which are not performed in this study.

Finally, the authors show that protein synthesis and ER Ca2+ defects seem to contribute to the defect but they do not discuss the relationship between the two defects. If the authors treat the Rab10 KD cells with both ionomycin and Leucine, do they get a full rescue? Or is one defect upstream of the other (e.g. can they see rescue of ER morphology upon Leucine treatment)? While this is not critical for the conclusions of the paper, several additional experiments could be performed to clarify their model, especially considering there is no clear model that explains how Rab10, protein synthesis, ER homeostasis, and Ca2+ are related to DCV (but not SV) exocytosis.

Reviewer #3 (Public Review):

In the submitted manuscript, Dong and colleagues set out to dissect the role of the Rab10 small GTPase on the intracellular trafficking and exocytosis of dense core vesicles (DCVs). While the authors have already shown that Rab3 plays a central role in the exocytosis of DVC in mammalian neurons, the roles of several other Rab-members have been identified genetically, but their precise mechanism of action in mammalian neurons remains unclear. In this study, the authors use a carefully designed and thoroughly executed series of experiments, including live-cell imaging, functional calcium-imaging, proteomics, and electron microscopy, to identify that DCV secretion upon Rab10 depletion in adult neurons is primarily a result of dysregulated protein synthesis and, to a lesser extent, disrupted intracellular calcium buffering. Given that the full deletion of Rab10 has a deleterious effect on neurons and that Rab10 has a major role in axonal development, the authors cautiously employed the knock-down strategy from 7 DIV, to focus on the functional impact of Rab10 in mature neurons. The experiments in this study were meticulously conducted, incorporating essential controls and thoughtful considerations, ensuring rigorous and comprehensive results.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

Dong et al here have studied the impact of the small Ras-like GTPase Rab10 on the exocytosis of dense core vesicles (DVC), which are important mediators of neuropeptide signaling in the brain. They use optical imaging to show that lentiviral depletion of Rab10 in mouse hippocampal neurons in culture independent of the established defects in neurite outgrowth hamper DCV exocytosis. They further demonstrate that such defects are paralleled by changes in ER morphology and defective ER-based calcium buffering as well as reduced ribosomal protein expression in Rab10-depleted neurons. Re-expression of Rab10 or supplementation of exogenous L-leucine to restore defective neuronal protein synthesis rescues impaired DCV secretion. Based on these results they propose that Rab10 regulates DCV release by maintaining ER calcium homeostasis and neuronal protein synthesis.

Strengths:

This work provides interesting and potentially important new insights into the connection between ER function and the regulated secretion of neuropeptides via DCVs. The authors combine advanced optical imaging with light and electron microscopy, biochemistry, and proteomics approaches to thoroughly assess the effects of Rab10 knockdown at the cellular level in primary neurons. The proteomic dataset provided may be valuable in facilitating future studies regarding Rab10 function. This work will thus be of interest to neuroscientists and cell biologists.

We appreciate the positive evaluation of our manuscript.

Weaknesses:

While the main conclusions of this study are comparably well supported by the data, I see three major weaknesses:

(1) For some of the data the statistical basis for analysis remains unclear. I.e. is the statistical assessment based on N= number of experiments or n = number of synapses, images, fields of view etc.? As the latter cannot be considered independent biological replicates, they should not form the basis of statistical testing.

This is an important point and we agree that multiple samples from the same biological replicate are not independent observations. We reanalyzed all nested data using a linear mixed model and indicated this in the Methods section and the relevant figure legends (Brunner et al., 2022). In brief, biological replicates (individual neuronal cultures) were used as a linear predictor. Outliers were identified and excluded using the ROUT method in GraphPad. A fixed linear regression model was then fitted to the data using the lm() function in R. A one-way anova (analysis of variance) was used to assess whether including the experimental group as a second linear predictor (formula = y ~ Group + Culture) statistically improved the fit of a model without group information (formula = y ~ 1 + Culture). Post-hoc analysis was performed using the emmeans() function with Tukey’s adjustment when more than two experimental groups were present. Importantly, our conclusions remain unchanged.

(2) As it stands the paper reports on three partially independent phenotypic observations, the causal interrelationship of which remains unclear. Based on prior studies (e.g. Mercan et al 2013 Mol Cell Biol; Graves et al JBC 1997) it is conceivable that defective ER-based calcium signaling and the observed reduction in protein synthesis are causally related. For example, ER calcium release is known to promote pS6K1 phosphorylation, a major upstream regulator of protein synthesis and ribosome biogenesis. Conversely, L-leucine supplementation is known to trigger calcium release from ER stores via IP3Rs. Given the reported impact of Rab10 on axonal transport of autophagosomes and, possibly, lysosomes via JIP3/4 or other mediators (see e.g. Cason and Holzbaur JCB 2023) and the fact that mTORC1, the alleged target of leucine supplementation, is located on lysosomes, which in turn form membrane contacts with the ER, it seems worth analyzing whether the various phenotypes observed are linked at the level of mTORC1 signaling.

This is great suggestion that could indeed further clarify the potential interplay between ER-based Ca2+ signaling and protein synthesis. To address this, we assessed the phosphorylation level of pS6K1 in control and Rab10 knockdown (KD) neurons with or without leucine treatment. These data are included in the new Figure 8—figure supplement 1 in the revised manuscript. Our results indicate that pS6K1 phosphorylation was not upregulated in Rab10 KD neurons, suggesting that the level of mTORC1 signaling is not different between wild-type or KD neurons. Furthermore, leucine treatment increased the pS6K1 phosphorylation level, as expected, but this effect was similar in both groups. Hence, we conclude that differences in mTORC1 signaling induced by Rab10 loss is not a major factor in the observed impairment in protein synthesis.

Rab10 depletion does not upregulate mTORC1 pathway.

(A)Typical immunoblot showing pS6K1 levels in each condition. (B) Quantification of relative pS6K1 levels in each condition. All Data are plotted as mean±s.e.m. (C) Control, Control + Leu: N = 2, n = 2, Rab10 KD, Rab10 KD + Leu: N = 2, n = 4.

(3) The claimed lack of effect of Rab10 depletion on SV exocytosis is solely based on very strong train stimulation with 200 Aps, a condition not very well suited to analyze defects in SV fusion. The conclusion that Rab10 loss does not impact SV fusion thus seems premature.

We agree that 200 APs stimulation might be too strong to detect specific effects on evoked synaptic vesicle release, although this stimulation pattern is an established pattern in hundreds of studies (Emperador-Melero et al., 2018; Granseth et al., 2006; Ivanova et al., 2021; Kwon and Chapman, 2011; Reshetniak et al., 2020). We have toned down our conclusions and clarified in the revised manuscript that Rab10 is dispensable for SV exocytosis evoked by intense stimulations. The corresponding statements in the text have been modified accordingly (p. 5, l. 98, 124) and in figure legend (p. 17, 490).

Reviewer #2 (Public Review):

Summary:

In this paper, the authors assess the function of Rab10 in dense core vesicle (DCV) exocytosis using RNAi and cultured neurons. The author provides evidence that their knockdown (KD) is effective and provides evidence that DCV is compromised. They also perform proteomic analysis to identify potential pathways that are affected upon KD of Rab10 that may be involved in DCV release. Upon focusing on ER morphology and protein synthesis, the authors conclude that defects in protein synthesis and ER Ca2+ homeostasis contributes to the DVC release defect upon Rab10 KD. The authors claim that Rab10 is not involved in synaptic vesicle (SV) release and membrane homeostasis in mature neurons.

Strengths:

The data related to Rab10's role in DCV release seems to be strong and carried out with rigor. While the paper lacks in vivo evidence that this gene is indeed involved in DCV in a living mammalian organism, I feel the cellular studies have value. The identification of ER defect in Rab10 manipulation is not truly novel but it is a good conformation of studies performed in other systems. The finding that DCV release defect and protein synthesis defect seen upon Rab10 KD can be significantly suppressed by Leucine supplementation is also a strength of this work.

We appreciate the positive evaluation of our manuscript.

Weaknesses:

The data showing Rab10 is NOT involved in SV exocytosis seems a bit weak to me. Since the proteomic analysis revealed so many proteins that are involved in SV exo/encodytosis to be affected upon Rab10, it is a bit strange that they didn't see an obvious defect. Perhaps this could have been because of the protocol that the authors used to trigger SV release (I am not an E-phys expert but perhaps this could have been a 'sledge-hammer' manipulation that may mask any subtle defects)? Perhaps the authors can claim that DCV is more sensitive to Rab10 KD than SV, but I am not sure whether the authors should make a strong claim about Rab10 not being important for SV exocytosis.

We agree that 200 APs stimulation might be too strong to see specific effects on evoked synaptic vesicle release, although this stimulation pattern is an established pattern in hundreds of studies. We have toned down our conclusions and clarified in the revised manuscript that Rab10 is dispensable for SV exocytosis evoked by intense stimulations. The corresponding statements in the text have been modified accordingly (p. 5, l. 98, 124) and in figure legend (p. 17, 490).

Also, the authors mention "Rab10 does not regulate membrane homeostasis in mature neurons" but I feel this is an overstatement. Since the authors only performed KD experiments, not knock-out (KO) experiments, I believe they should not make any conclusion about it not being required, especially since there is some level of Rab10 present in their cells. If they want to make these claims, I believe the authors will need to perform conditional KO experiments, which are not performed in this study.

This is a valid point. We have changed the statement to “membrane homeostasis in mature neurons was unaffected by Rab10 knockdown” (p. 13, l.376-377).

Finally, the authors show that protein synthesis and ER Ca2+ defects seem to contribute to the defect but they do not discuss the relationship between the two defects. If the authors treat the Rab10 KD cells with both ionomycin and Leucine, do they get a full rescue? Or is one defect upstream of the other (e.g. can they see rescue of ER morphology upon Leucine treatment)? While this is not critical for the conclusions of the paper, several additional experiments could be performed to clarify their model, especially considering there is no clear model that explains how Rab10, protein synthesis, ER homeostasis, and Ca2+ are related to DCV (but not SV) exocytosis.

This is an important point and a great suggestion. We have now tested the rescue effects of leucine treatment on ER morphology, as suggested. These data are included in the new Figure 8—figure supplement 2 in the revised manuscript. Our results indicate that the same dose of leucine that rescues DCV fusion and protein translation failed to rescue ER morphology. Hence, the defects in ER morphology appear to be independent of the impaired protein translation.

Leucine supplementation does not rescue ER morphological deficiency in Rab10 KD neurons.

(A) Typical examples showing the KDEL signals in each condition. (B) Quantification of RTN4 intensity in MAP2-positive dendrites. (C) The ratio of neuritic to somatic RTN4 intensity (N/S). All Data are plotted as mean±s.e.m. (B, C) Control: N = 3, n = 10; Rab10 KD: N = 3, n = 11; Rab10 KD + Leu: N = 3; n = 11. A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, ns = not significant.

Reviewer #3 (Public Review):

In the submitted manuscript, Dong and colleagues set out to dissect the role of the Rab10 small GTPase on the intracellular trafficking and exocytosis of dense core vesicles (DCVs). While the authors have already shown that Rab3 plays a central role in the exocytosis of DVC in mammalian neurons, the roles of several other Rab-members have been identified genetically, but their precise mechanism of action in mammalian neurons remains unclear. In this study, the authors use a carefully designed and thoroughly executed series of experiments, including live-cell imaging, functional calcium-imaging, proteomics, and electron microscopy, to identify that DCV secretion upon Rab10 depletion in adult neurons is primarily a result of dysregulated protein synthesis and, to a lesser extent, disrupted intracellular calcium buffering. Given that the full deletion of Rab10 has a deleterious effect on neurons and that Rab10 has a major role in axonal development, the authors cautiously employed the knock-down strategy from 7 DIV, to focus on the functional impact of Rab10 in mature neurons. The experiments in this study were meticulously conducted, incorporating essential controls and thoughtful considerations, ensuring rigorous and comprehensive results.

We are grateful for the positive evaluation of our manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

The work by Dong et al provides interesting and potentially important new insights into the connection between ER function and the regulated secretion of neuropeptides via DCVs. I suggest that the authors address the following points experimentally to increase the impact of this potentially important study.

Major points:

(1) As alluded to above, for some of the data the statistical basis for analysis remains unclear (examples are Figures 1C-F, J,K; Figure 2 1B-D,I-K; Figure 2 - Supplement 1D-F; Figure 2 - Supplement 2J,K, etc). I.e. is the statistical assessment based on N = number of experiments or n = number of synapses, images, fields of view etc.? As the latter cannot be considered independent biological replicates, they should not form the basis of statistical testing. The Ms misses also misses a dedicated paragraph on statistics in the methods section.

See reply to reviewer 1 above. We fully agree and solved this point.

(2) A main weakness of the paper is the missing connection between neuronal protein synthesis, and the observed structural and signaling defects at the level of the ER. I suggest that the authors analyze mTORC1 signaling in Rab10 depleted neurons and under rescue conditions (+Leu or re-expression of Rab10) as ribosome biogenesis is a major downstream target of mTORC1 and mTORC1 activity is related to lysosome position, which may be affected upon rab10 loss -either directly or via effects on the ER that forms tight contacts with lysosomes.

See reply to reviewer 1 above. We agreed and followed up experimentally.

(3) Related to the above: Does overexpression of SERCA2 restore normal DCV exocytosis in Rab10-depleted neurons? This would help to distinguish whether calcium storage and release at the level of the ER indeed contribute to the exocytosis defect.

This is an important point and a great suggestion. We have now tested the rescue effects of overexpression of SERCA2 on DCV fusion. These data are included in the new Figure 8—figure supplement 3 in the revised manuscript. SERCA2 OE failed to rescue the DCV fusion defects in Rab10 KD neurons.

Overexpression of SERCA2 does not rescue DCV fusion deficits in Rab10 KD neurons.

(A) Typical examples showing the SERCA2 signals in each condition. (B) Cumulative plot of DCV fusion events per cell. (C) Summary graph of DCV fusion events per cell. (A) Total number of DCVs (total pool) per neuron, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion. (B) Fraction of NPY-pHluorin-labeled DCVs fusing during stimulation. All Data are plotted as mean±s.e.m. (C-E) Control: N = 2, n = 10; Rab10 KD: N = 2, n = 13; SERCA2 OE: N = 2; n = 15. A one-way ANOVA tested the significance of adding experimental group as a predictor. *** = p<0.001, ** = p<0.01, ns = not significant.

(4) The claimed lack of effect of Rab10 depletion on SV exocytosis is solely based on very strong train stimulation with 200 Aps, a condition not very well suited to analyze defects in SV fusion. The conclusion that Rab10 loss does not impact SV fusion thus seems premature. The authors should conduct additional experiments under conditions of single or few Aps (e.g. 4 or 10 Aps) to really assess whether or not Rab10 depletion alters SV exocytosis at the level of pHluorin analysis in cultured neurons.

See reply to reviewer 2 above. Agreed to and made textual adjustments to solve this

(5) Related to the above: I am puzzled by the data shown in Figure 1H-J: From the pHluorin traces shown I would estimate a tau value of about 20-30 s (e.g. decay to 1/e = 37% of the peak value). The bar graph in Figure 1K claims 3-4 s, clearly clashing with the data shown. Were these experiments conducted at RT (where expected tau values are in the range of 30s) or at 37{degree sign}C (one would expect taus of around 10 s in this case for Syp-pH)? I ask the authors to carefully check and possibly re-analyze their datasets.

This is indeed a mistake. We thank the reviewer for flagging this miscalculation. Our original Matlab script used for calculating the tau value contained an error and the datasets were normalized twice by mistake. We now reanalyzed the data and the corresponding figures and texts have been updated. Our conclusion that Rab10 KD does not affect SV endocytosis remains unchanged since the difference in tau between the control (28.5 s) and Rab10 KD (32.8 s) suffered from the same systematic error and were/are not significantly different.

(6) How many times was the proteomics experiment shown in Figure 3 conducted? I noticed that the data in panel H missed statistical analysis and error bars. Given the typical variation in these experiments, I suggest to only include data for proteins identified in at least 3 out of 4 experimental replicates.

We agree that this information has not been clear. We have now explained replication in the Methods section (p. 42, l. 879-885). In brief, the proteomics experiment presented in Fig 3 was conducted with two independent cultures (‘biological replicates’), hence, formally only two independent observations. For each biological replicate, we performed four technical replicates. For our analysis, we only included peptides that were consistently detected across all samples (not only three as this reviewer suggests). Proteins in Panel H are ER-related proteins that are significantly different from control neurons with an adjusted FDR ≤ 0.01 and Log2 fold change ≥ 0.56. The primary purpose of our proteomics experiments was to generate hypotheses and guide subsequent experiments and the main findings were corroborated by other experiments presented in the manuscript.

Minor:

(7) Figure 2 - supplement 3 and Figure 4 - supplement 3 are only mentioned in the discussion. The authors should consider referring to these data in the results section.

This is a valid point. We have now added a new statement “Moreover, only 10% of DCVs co-transport with Rab10” in the Results (p. 6-7, l. 162-164).

(8) Where is the pHluorin data shown in Figure 1 bleach-corrected? If so, this should be stated somewhere in the Ms. Moreover, the timing of the NH4Cl pulse should be indicated in the scheme in panel I.

We thank the reviewer for pointing these omissions out. We have now included information about the timing of NH4Cl pulse in panel I. We did not do bleach-correction for the pHluorin data shown in Figure 1. It has been shown that pHluorin is very stable with a bleaching rate in the alkaline state of 0.06% per second and 0.0024% per second in the quenched state (Balaji and Ryan, 2007). Indeed, we did not observe obvious photobleaching in the first 30s during our imaging as indicated by the average trace of pHluorin intensity in panel I.

(9) Page 3/ lines 59-60: "...strongest inhibition of neuropeptide accumulation...". What is probably meant is "...strongest inhibition of neuropeptide release".

We agree this statement is unclear. Sasidharan et al used a coelomocyte uptake assay as an indirect readout for DCV release. The ‘strongest inhibition of neuropeptide accumulation’ in coelomocytes in Rab10 mutant indicates DCV fusion deficits. We have now replaced the text with “Rab10 deficiency produces the strongest inhibition of neuropeptide release in C. elegans” to make it more clear.

Reviewer #3 (Recommendations For The Authors):

I strongly recommend the publishing of this study as a VOR with minor comments directed to the authors.

(1) In Figure 4, the authors should include examples of tubular ER at the synapse, especially as this is an interesting point discussed in ln 226-229. Are there noticeable changes in the ER-mitochondria contacts at the synaptic boutons?

We agree that examples of tubular ER at the synapse would improve the manuscript. We have now replaced the Figure 4A with such examples. We found it challenging to quantify ER-mitochondria contacts based on the electron microscopy (EM) images we currently have. The ER-mitochondria contact sites are quite rare in the cross-sections of our samples, making it difficult to perform a reliable quantitative analysis.

(2) The limited impairment of calcium-ion homeostasis in Rab10 KD neurons is very interesting. Would the overexpression of Rab10T23N mimic the effect of a KD scenario? Is there a separation of function for Rab10 in calcium homeostasis vs. the regulation of protein synthesis?

This is an interesting possibility. We tested this and expressed Rab10T23N in a new series of experiments. These data are presented as a new Figure 5 in the revised manuscript (p. 29). We observed that Ca2+ refilling after caffeine treatment was delayed to a similar extent in Rab10T23N-expressing and Rab10 KD neurons. While impaired Ca2+ homeostasis may affect protein synthesis through ER stress or mTORC1 activation, our findings indicate otherwise in Rab10 KD neurons. First, ATF4 levels, a marker of ER stress, were unaffected in Rab10 KD neurons. This indicates that any ER stress present is minimal or insufficient to significantly impact protein synthesis through this pathway. Second, we did not observe significant changes in mTORC1 activation in Rab10 KD neurons as indicated by a normal pS6K1 phosphorylation (see above). Based on these observations, we conclude that Rab10's roles in calcium homeostasis and protein synthesis are most likely separate.

(3) The authors indicate that the internal release of calcium ions from the ER has no effect on DCV trafficking and fusion without showing the data. It is important to include this data as the major impact of the study is the dissecting of the calcium effects in mammalian neurons from the previous studies in invertebrates.

We agree this is an important aspect in our reasoning. We are submitting the related manuscript on internal calcium stores to BioRVix. The link will be added to the consolidated version of our manuscript

(4) The distinction between Rab3 and Rab10 co-trafficking on DCVs should be reported in the Results (currently, Figure 2 - supplement 3 is only mentioned in the Discussion) as it helps to understand the effects on DCV fusion.

We agree. We now added a new statement “Moreover, only 10% of DCVs co-transport with Rab10” in the Results (p. 6, l. 162-163).

Reference:

Balaji, J., Ryan, T.A., 2007. Single-vesicle imaging reveals that synaptic vesicle exocytosis and endocytosis are coupled by a single stochastic mode. Proceedings of the National Academy of Sciences 104, 20576–20581. https://doi.org/10.1073/pnas.0707574105

Brunner, J.W., Lammertse, H.C.A., Berkel, A.A. van, Koopmans, F., Li, K.W., Smit, A.B., Toonen, R.F., Verhage, M., Sluis, S. van der, 2022. Power and optimal study design in iPSC-based brain disease modelling. Molecular Psychiatry 28, 1545. https://doi.org/10.1038/s41380-022-01866-3

Emperador-Melero, J., Huson, V., van Weering, J., Bollmann, C., Fischer von Mollard, G., Toonen, R.F., Verhage, M., 2018. Vti1a/b regulate synaptic vesicle and dense core vesicle secretion via protein sorting at the Golgi. Nat Commun 9, 3421. https://doi.org/10.1038/s41467-018-05699-z

Granseth, B., Odermatt, B., Royle, S.J., Lagnado, L., 2006. Clathrin-Mediated Endocytosis Is the Dominant Mechanism of Vesicle Retrieval at Hippocampal Synapses. Neuron 51, 773–786. https://doi.org/10.1016/j.neuron.2006.08.029

Ivanova, D., Dobson, K.L., Gajbhiye, A., Davenport, E.C., Hacker, D., Ultanir, S.K., Trost, M., Cousin, M.A., 2021. Control of synaptic vesicle release probability via VAMP4 targeting to endolysosomes. Science Advances 7, eabf3873. https://doi.org/10.1126/sciadv.abf3873

Kwon, S.E., Chapman, E.R., 2011. Synaptophysin Regulates the Kinetics of Synaptic Vesicle Endocytosis in Central Neurons. Neuron 70, 847–854. https://doi.org/10.1016/j.neuron.2011.04.001

Reshetniak, S., Fernández-Busnadiego, R., Müller, M., Rizzoli, S.O., Tetzlaff, C., 2020. Quantitative Synaptic Biology: A Perspective on Techniques, Numbers and Expectations. International Journal of Molecular Sciences 21, 7298. https://doi.org/10.3390/ijms21197298

Version 2

Abstract

Abstract

Dense core vesicles (DCVs) transport and release various neuropeptides and neurotrophins that control diverse brain functions, but the DCV secretory pathway remains poorly understood. Here, we tested a prediction emerging from invertebrate studies about the crucial role of the intracellular trafficking GTPase Rab10, by assessing DCV exocytosis at single-cell resolution upon acute Rab10 depletion in mature mouse hippocampal neurons, to circumvent potential confounding effects of Rab10’s established role in neurite outgrowth. We observed a significant inhibition of DCV exocytosis in Rab10-depleted neurons, whereas synaptic vesicle exocytosis was unaffected. However, rather than a direct involvement in DCV trafficking, this effect was attributed to two ER-dependent processes, ER-regulated intracellular Ca2+ dynamics and protein synthesis. Gene ontology analysis of differentially expressed proteins upon Rab10 depletion identified substantial alterations in synaptic and ER/ribosomal proteins, including the Ca2+-pump SERCA2. In addition, ER morphology and dynamics were altered, ER Ca2+ levels were depleted and Ca2+ homeostasis was impaired in Rab10-depleted neurons. However, Ca2+ entry using a Ca2+ ionophore still triggered less DCV exocytosis. Instead, leucine supplementation, which enhances protein synthesis, largely rescued DCV exocytosis deficiency. We conclude that Rab10 is required for neuropeptide release by maintaining Ca2+ dynamics and regulating protein synthesis. Furthermore, DCV exocytosis appeared more dependent on (acute) protein synthesis than synaptic vesicle exocytosis.

Introduction

Dense core vesicles (DCVs) transport and release neuromodulators (neuropeptides, neurotrophic factors, and catecholamines) that play crucial roles in modulating diverse brain functions, including sleep, mood, memory, and learning (e.g., Cawley et al., 2016; Malva et al., 2012; Poo, 2001; Salio et al., 2006). Deficits in neuropeptide signaling pathways have been associated with several human disorders and diseases, including anxiety, depression, and obesity (Beck, 2000; Sah and Geracioti, 2013; Barde et al., 2022). Neuropeptides are synthesized in the endoplasmic reticulum (ER) and subsequently packaged into immature DCVs in the Golgi complex (Tooze et al., 2001). DCVs are transported along the neurites and undergo activity-dependent membrane fusion with the plasma membrane to release their content (Farina et al., 2015; Heidelberger et al., 1994; Nassal et al., 2022; Thomas et al., 1993; van de Bospoort et al., 2012). Despite the critical role of neuropeptides in brain functions, the regulatory mechanisms governing their secretion are not fully understood. We have shown previously that Rab3 is an essential regulator in the last steps of the DCV secretory pathway in mammalian neurons (Persoon et al., 2019). However, studies in invertebrates have also implicated other Rab proteins, including Rab2, Rab5, and Rab10, in the DCV secretory pathway: (Ailion et al., 2014; Azouz et al., 2014; Edwards et al., 2009; Hannemann et al., 2012; Lund et al., 2020; Sasidharan et al., 2012; Sumakovic et al., 2009).

Among these Rab proteins, Rab10 deficiency produces the strongest inhibition of neuropeptide release in C. elegans (Sasidharan et al., 2012). With its subcellular localization in many vesicular organelles, such as plasmalemmal precursor vesicles (PPVs), GLUT4 transport vesicles, and recycling endosomes, Rab10 regulates various aspects of intracellular membrane trafficking, including vesicle formation, transport, and fusion (Chen et al., 2006; Larance et al., 2005; Mîinea et al., 2005; Sano et al., 2007; Taylor et al., 2015). Rab10 deficiency leads to deficits in these pathways, resulting in impaired neuronal outgrowth and disrupted retrograde axonal transport of signaling factors (Lazo and Schiavo, 2023). How these deficits relate to the strong inhibition of DCV exocytosis remains unknown.

Deletion of Rab10 expression or inhibiting its functions by overexpression of an inactive mutant (Rab10T23N) leads to abnormal ER morphology (English and Voeltz, 2013; Lv et al., 2015; Shih and Hsueh, 2016). Since the ER is a crucial organelle involved in protein synthesis, Ca2+ buffering, and lipid metabolism, alterations in its structure and function can directly or indirectly affect neuronal secretory pathways. Indeed, several studies have suggested the roles of the ER in the DCV pathway, such as the involvement of ER stress and lipid levels in DCV production in C. elegans (Laurent et al., 2018; Valadas et al., 2018), and the roles of ER Ca2+ as an internal Ca2+ source regulating somatodendritic dopamine release in mouse substantia nigra neurons (Patel et al., 2009). Additionally, Rab10 mutations, altered expression levels, or phosphorylation states are firmly associated with CNS disorders (Agola et al., 2011; Cheng et al., 2005; Ishida et al., 2016; Kiral et al., 2018). Hence, the strong inhibition of neuropeptide accumulation in coelomocytes in nematode Rab10 mutants may be a direct or indirect effect of Rab10 loss of function on DCV exocytosis and Rab10-dependent disease processes may (in part) be explained by DCV exocytosis impairment.

In the present study, we investigated the involvement of Rab10 in regulated secretion of neuropeptides in mammalian neurons. We directly assessed DCV exocytosis at single-cell resolution in mouse hippocampal neurons and confirmed that Rab10 is a crucial regulator of DCV exocytosis also in mammalian systems, while not affecting synaptic vesicle (SV) exocytosis. However, instead of having a direct role in the DCV secretory pathway, we observed that Rab10 is involved in DCV exocytosis through ER-dependent processes, especially reduced ER Ca2+ homeostasis and impaired global protein synthesis. Supplementation with leucine known to boost protein synthesis, rescued the defects in DCV exocytosis. We therefore conclude that Rab10 plays a central role in regulating DCV exocytosis by maintaining Ca2+ homeostasis and protein synthesis.

Results

Rab10 regulates neuronal outgrowth but is dispensable for synaptogenesis and SV exocytosis under intense stimulation

To study the role of Rab10 in regulated secretion, we utilized a knockdown strategy since complete knockout of Rab10 results in lethality at cell and organismal levels (Lv et al., 2015). We selected two specific shRNA sequences, shRNA#9, and shRNA#11, to deplete Rab10 expression, and a scrambled sequence as control. In mouse cortical neurons infected with either shRNA#9 or shRNA#11 at day in vitro 0 (DIV0), we observed a 75-95% decrease in Rab10 expression at DIV14 (Figure 1A). Previous studies have shown that Rab10 regulates neuronal outgrowth (Fukuda, 2008; Hutagalung and Novick, 2011). Consistent with these findings, we observed a significantly reduced dendrite and axon length in neurons infected with shRNA#9 at DIV0 (Figure 1B-D, 1G).

To test the effects of Rab10 depletion on synaptogenesis, synaptic vesicles (SVs) were quantified using the endogenous marker synaptophysin-1 (Syp1). Syp1 staining exhibited a punctate distribution at DIV14, indicating the accumulation of SVs in boutons/synapses, and no changes in the number of puncta per µm neurite or the intensity of Syp1 puncta (Figure 1E-F). These data confirm that Rab10 regulates neurite outgrowth, but we found no evidence for a role in synaptogenesis.

To test whether Rab10 depletion affects SV exocytosis, hippocampal neurons were infected with the SV exocytosis reporter Synaptophysin-pHluorin (SypHy; Figure 1H; Granseth et al., 2006). SV exocytosis was triggered by high-frequency electrical stimulation (HFS, 5s 40-Hz). The total vesicle pool was measured by briefly superfusing Tyrode’s solution containing 50 mM NH4Cl. The fraction of fused SVs, determined by the ratio of SypHy intensity upon HFS to the maximum intensity upon NH4Cl superfusion, was comparable in the two groups (Figure 1I-J). In addition, SV endocytosis, measured by the fluorescence decay of SypHy after HFS, was unaffected by Rab10 depletion (Figure 1K). Therefore, we conclude that Rab10 is dispensable for SV exocytosis under intense stimulation.

Rab10 is required for neuronal outgrowth but dispensable for SV exocytosis evoked by intense stimulations.

(A) Representative immunoblotting showing knockdown and rescue of Rab10 expression in cultured primary neurons infected with shRNA against Rab10 or rescue constructs (upper) and quantification of Rab10 levels (bottom).

(B) Example images of control or Rab10 KD hippocampal neurons (DIV14) stained for the dendrite marker MAP2 (blue), the synapse marker Syp1 (red), and the axonal marker SMI312 (green). Scale bar: 50 μm (upper) and 10 μm (bottom).

(C) Quantification of the dendritic length (MAP2).

(D) Quantification of the axonal length (SMI312).

(E) Quantification of Syp1 intensity per synapse per neuron.

(F) Quantification of the Syp1-positive synapse density in MAP2-positive dendrites.

(G) Sholl analysis showing the mean number of dendritic branches against the distance from the soma.

(H) Example neurons infected with the SV fusion marker SypHy (upper), typical kymographs of neurites showing SypHy intensity increase during stimulation and upon NH4Cl superfusion (bottom).

(I) The average signal SypHy from active synapses, normalized from baseline to maximum fluorescence upon NH4Cl superfusion.

(J) SV exocytosis determined as the ratio of the maximum SypHy intensity during stimulation to the maximum during NH4Cl stimulation.

(K) SV endocytosis determined as the SypHy signal decay time constant τ in the 60 s after field stimulation.

All Data are plotted as mean±s.e.m. (A) N = 4, n = 4, one sample t-test. (C-G) Control: N = 3, n = 35; ShRNA#9: N = 3, n = 32. (J, K) Control: N = 3, n = 56; ShRNA#9: N = 3, n = 56. (C-F, J, K) A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, ** = p<0.01, ns = not significant.

Rab10 is a major regulator of DCV exocytosis

To investigate the role of Rab10 in neuropeptide release, we depleted Rab10 levels at DIV0, expressed the DCV exocytosis reporter NPY-pHluorin at DIV9-10, and performed live-cell imaging at DIV14. Our previous studies have demonstrated that NPY-pHluorin almost exclusively localizes to DCVs as indicated by its strong co-localization with endogenous markers of DCVs, such as BDNF and the chromogranins CHGA and CHGB (Arora et al., 2017; Dominguez et al., 2018; Persoon et al., 2019). To achieve single-vesicle resolution analysis of DCV exocytosis, we used single cultured hippocampal neurons on glial micro-islands. Neurons were stimulated by 16 trains of 50 action potentials (APs) at 50 Hz, a protocol known to trigger robust DCV exocytosis (Balkowiec and Katz, 2002; Emperador-Melero et al., 2018; Gärtner and Staiger, 2002; Hartmann et al., 2001; Moro et al., 2021; Persoon et al., 2019). Fusion events were detected as a rapid appearance of fluorescent puncta (Figure 2—figure supplement 1A-B). DCV exocytosis in Rab10 KD neurons was significantly reduced by 60% compared to control neurons (Figure 2—figure supplement 1C-D). Furthermore, the total number of DCVs was reduced by 30% in Rab10 KD neurons (Figure 2—figure supplement 1E). The fusion fraction, which represents the number of DCV fusion events relative to the remaining DCV pool after stimulation, was also significantly decreased by 50% in Rab10 KD neurons (Figure 2—figure supplement 1F). Overexpression of wild-type, knockdown-resistant Rab10 rescued DCV exocytosis deficits in Rab10 KD neurons (Figure 2—figure supplement 1C-F).

Rab10 is a major regulator of DCV exocytosis.

(A) Example images of control and Rab10 KD hippocampal neurons (DIV14) stained for MAP2 (blue), Syp1 (red) and SMI312 (green). Scale bar: 30 μm.

(B) Quantification of the dendritic length (MAP2).

(C) Quantification of the axonal length (SMI312).

(D) Quantification of the Syp1-positive synapse density in MAP2-positive dendrites.

(E) Sholl analysis showing the mean number of dendritic branches against the distance from the soma.

(F) Schematic representation of DCV fusion assay. DCVs are labeled with NPY-pHluorin, and neurons are stimulated with one train of 16 bursts of 50 APs at 50 Hz (light blue bars).

(G) Representative neurons during electrical stimulation superimposed with NPY-pHluorin fusion events (green dots). Scale bar: 5 μm.

(H) Cumulative plot of DCV fusion events per cell. Light blue bars represent the stimulation trains.

(I) Summary graph of DCV fusion events per cell.

(J) The total number of DCVs (total pool) of neurons analyzed in Figure 2H, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

(K) Fraction of NPY-pHluorin-labeled DCV fusing during stimulation.

All Data are plotted as mean±s.e.m. (B-D) Control: N = 3, n = 31; ShRNA#9: N = 3, N = 28; ShRNA#11: N = 3, n = 31. (I-K) Control: N = 4, n = 36; shRNA#9: N = 4, N = 37; shRNA#11: N = 4, n = 30; Rescue: N = 4, n = 34. A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, *** = p<0.001, ** = p<0.01, * = p <0.05, ns = not significant.

Rab10 depletion at DIV0 impedes DCV fusion.

(A) Schematic representation of DCV fusion assay. DCVs are labeled with NPY-pHluorin, and neurons are stimulated with one train of 16 bursts of 50 APs at 50 Hz (light blue bars).

(B) Representative neurons during electrode stimulation superimposed with NPY-pHluorin fusion events (green dots). Scale bar: 10 μm.

(C) Cumulative plot of DCV fusion events per cell.

(D) Summary graph of DCV fusion events per cell.

(E) Total number of DCVs (total pool) of neurons, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

(F) Fraction of NPY-pHluorin-labeled DCV fusing during stimulation.

All Data are plotted as mean±s.e.m. (D-F) Control: N = 3, n = 26; Rab10 KD: N = 3, n = 47; Rescue: N = 3, n = 22. (D-F) A one-way ANOVA tested the significance of adding experimental group as a predictor. *** = p<0.001, ** = p<0.01, * = p <0.05.

Rab10 depletion does not affect DCV transport or cargo loading.

(A) Representative kymographs illustrating the transport of NPY-mCherry labeled DCVs in control and Rab10 KD neurons.

(B) Quantification of average velocity (µm/s) of control and Rab10 KD neurons.

(C) Quantification of average distance moved from the start (µm) of control and Rab10 KD neurons.

(D) Histogram of average velocity (µm/s) of control and Rab10 KD neurons.

(E) Histogram of average distance moved from the start (µm) of control and Rab10 KD neurons.

(F) Typical neurite expressing NPY-pHluorin during baseline (b), and during stimulation (s).

(G) Average traces of NPY-pHluorin fusion events aligned at the moment of fusion (0 seconds).

(H) Quantification of NPY-pHluorin baseline fluorescence before stimulation.

(I) Quantification of average NPY-pHluorin fusion intensity per cell.

All Data are plotted as mean±s.e.m. (B, C) Control: N = 3, n = 18; Rab10 KD: N = 3, n = 17. (H, I) Control: N = 3, n = 37; Rab10 KD: N = 3, n = 35. A one-way ANOVA tested the significance of adding experimental group as a predictor. ns = not significant.

Rab10 does not typically co-transport together with DCVs.

(A) Representative kymographs of neurons co-infected with Rab10-GFP and NPY-mCherry.

(B) Percentage moving DCVs that co-transport with Rab10. Data are plotted as mean±s.e.m (N = 3, n = 22).

Data are plotted as mean±s.e.m.

To overcome the potential confounding effects of impaired neurite outgrowth (Figure 1) and reduced total DCV pool (Figure 2—figure supplement 1) upon Rab10 depletion starting at DIV0, we adopted a more acute approach to interfere with Rab10 function, and late enough not to affect neuronal morphology and the total DCV pool. Neurons were transfected with shRNA against Rab10 at DIV7, fixed at DIV14, and stained with markers for dendrites (MAP2), axons (SMI312), and SVs (Syp1). Rab10 expression was reduced by 70% after 7 days of infection (Appendix 1—figure 3). No significant alterations in total dendrite length, axon length (Figure 2A-C, 2E), or synapse density (Figure 2D) were observed in Rab10 KD neurons under these conditions. Therefore, to eliminate confounding effects on morphological parameters, we reevaluated DCV exocytosis and all further experiments in neurons infected with shRNA at DIV7. DCV exocytosis in Rab10 KD neurons remained significantly lower by 50% compared to control neurons (Figure 2F-I). The fusion fraction was also significantly reduced by 65% in Rab10 KD neurons (Figure 2K). Overexpression of wild-type Rab10 rescued DCV exocytosis deficits. No significant differences in the total number of DCVs (Figure 2J), DCV transport (Figure 2—figure supplement 2A-2B, 2E), or cargo loading (Figure 2—figure supplement 2F-K) were observed. Moreover, only 10% of DCVs co-transport with Rab10 (Figure 2—figure supplement 3). Thus, these data indicate that Rab10 depletion specifically inhibits activity-dependent neuropeptide release in hippocampal neurons, without effects on DCV biogenesis, cargo loading, and transport and independent of Rab10’s established role in neuronal outgrowth.

Depletion of Rab10 leads to dysregulation of proteins enriched in presynaptic transmission and cytosolic translation.

(A) Volcano plots showing significantly dysregulated proteins in Rab10 depleted neurons.

(B) GO enrichment analysis of functional pathways of the significant hits with ClueGO. Shown are the Bonferroni corrected p-values.

(C) GO enrichment analysis of subcellular localization of the significant hits with ClueGO. Shown are the Bonferroni corrected p-values.

(D) Sunburst plot showing the annotation in synaptic function of the altered proteins in Rab10 depleted neurons.

(E) Sunburst plot showing the annotation in synaptic location of the altered proteins in Rab10 depleted neurons.

(F) Log2 fold changes of synaptic proteins within SynGO terms. Downregulated proteins are shown in blue and upregulated proteins are shown in black.

(G) Examples of proteins that are significantly affected by Rab10 depletion grouped by their subcellular localization. Heat maps represent the degree of up-or downregulation.

(H) Selective MS data analysis of ER-related proteins in Rab10 KD neurons. Bars show the fold change of the indicated peptides compared to the control.

Proteins involved in synaptic transmission and translation are severely dysregulated upon Rab10 depletion

To comprehensively investigate Rab10 function in mature neurons, mass spectrometry proteomics was performed on Rab10 KD and control neurons at DIV14. A total of approximately 5400 unique proteins were identified and quantified in two biological replicates. The complete list of proteins quantified in this study is presented in Appendix 1-figure 3. Only differentially expressed proteins detected with high confidence characterized by a log2(fold change) > 0.56, and q-value < 0.01, were included in the subsequent analysis. Among the dysregulated proteins, 71% were upregulated, while 29% were downregulated, resulting in a significant dysregulation of 19% of the total protein pool in Rab10 KD neurons. These data indicate that Rab10 depletion leads to major neuronal proteome dysregulation within 7 days after initiating knockdown (Figure 3A).

To gain insights into the functional consequences of Rab10 depletion, we performed Gene Ontology (GO) analysis using ClueGO (Bindea et al., 2009). This analysis revealed that biological processes related to chemical synaptic transmission were notably affected by Rab10 depletion (Figure 3B). In addition, several biological processes related to protein synthesis, such as cytoplasmic translation and ribosomal large subunit biogenesis, were among the top 5 terms with the lowest p-values (Figure 3B). Subcellular localization analysis of these dysregulated proteins upon Rab10 depletion showed that cytosolic ribosomal proteins were the most significantly affected, followed by dendritic proteins (Figure 3C).

Further characterization of the dysregulated synaptic proteins was performed using SynGO (Koopmans et al., 2019). Among the 391 significantly dysregulated proteins annotated in SynGO, 205 were classified as presynaptic proteins and 237 as postsynaptic proteins. GO enrichment analysis revealed that biological processes in metabolism were most dysregulated upon Rab10 depletion. Among them, both pre and postsynaptic translation were significantly dysregulated (Figure 3D). Consistent with the ClueGO analysis, SynGO highlighted a significant dysregulation of presynaptic (34 of 391) and postsynaptic ribosomal proteins (44 of 391), supporting the involvement of Rab10 in the regulation of neuronal protein synthesis (Figure 3E). Interestingly, all dysregulated ribosomal proteins were upregulated upon Rab10 depletion, which contrasts with the mostly downregulated expression observed in the other classes of proteins (such as synaptic and cytoskeletal proteins) (Figure 3F-G). Taken together, GO analysis with both ClueGO and SynGO indicates a dysfunction of protein translation in Rab10 KD neurons.

Loss of Rab10 expression has been associated with altered ER morphology in mouse embryonic cells (Lv et al., 2015), which may explain the selective upregulation of proteins involved in ribosome function in our proteomics data. Indeed, ER proteins were dysregulated substantially upon Rab10 depletion. Specifically, several rough ER proteins showed differential regulation, with SEC61A1 being upregulated, SEC61G being downregulated, and CLIMP remaining unchanged (Figure 3H). Most tubular ER proteins, such as RTN3/4 and VAPB, were robustly decreased. Interestingly, one of the ER membrane Ca2+ channels, SERCA2, showed a 50% reduction upon Rab10 depletion (Figure 3H).

Taken together, these analyses reveal that Rab10 depletion leads to major changes in protein expression, especially synaptic and ribosomal proteins, the latter all upregulated in Rab10-depleted neurons which suggests that protein synthesis is dysregulated, potentially due to altered ER function.

Rab10 regulates ER morphology and ribosomal protein levels

Given the substantial dysregulation of synapse and ribosome/ER proteins, we investigated synapses and ER further using electron microscopy (Figure 4). These analyses revealed an apparently normal synapse morphology in Rab10 depleted neurons with many SVs clustered at the active zone, while DCVs were sparsely distributed along neurites and near the active zone (Figure 4A). The length of the active zone (AZ) and postsynaptic density (PSD) were both decreased by 10% upon Rab10 depletion (Figure 4C-D). However, other parameters of synaptic ultrastructure, such as the diameter of SVs or DCVs, and the number of SVs per synapse, were unchanged in Rab10 KD neurons (Figure 4E-H). Hence, despite substantial dysregulation of synaptic proteins, overall synapse morphology was hardly affected.

Rab10 regulates synapse size and ER morphology.

(A) Representative EM pictures showing the ultrastructure of synapses. Scale bar 100 nm. Synaptic ER is indicated by red dotted lines.

(B) Representative EM pictures showing the ultrastructure of soma. Rough ER (rER) is indicated by red dotted lines. M: mitochondrion, G: Golgi. Scale bar: 100 nm.

(C) Quantification of the length of active zone and postsynaptic density (PSD).

(D) Quantification of the length of postsynaptic density (PSD).

(E) Quantification of SV number per synapse and SV diameter.

(F) Quantification of SV diameter.

(G) Quantification of DCV diameter.

(H) Frequency distribution of DCVs by diameter.

(I) Quantification of the diameter of rough ER (rER).

Data are plotted with Superplot (Figure. 4C-G, I), where averages from three independent cultures are shown as large circles and single observations are shown as dots. Horizontal lines represent the means of the averages from three weeks. Data from different cultures are grouped with different colors. (C-D) Control: N = 3, n = 184; shRNA#9: N = 3, n = 187. (E) Control: N = 3, n = 189; shRNA#9: N = 3, n = 188. (F) Control: N = 3, n = 1770; shRNA#9: N = 3, n = 1803. (G) Control: N = 3, n = 137; shRNA#9: N = 3, n = 122. (I) Control: N = 3, n = 63; shRNA#9: N = 3, n = 64. (C-G, I) linear mixed model analysis. *** = p<0.001, * = p <0.05, ns = not significant.

Altered ER morphology in Rab10 KD neurons.

(A) Example images of control or Rab10 KD hippocampal neurons (DIV14) stained for the dendrite marker MAP2 (green), two ER markers KDEL (red) and RTN4 (magenta). Scale bar: 50 μm.

(B) Quantification of RTN4 intensity in MAP2-positive dendrites.

(C) Quantification of KDEL intensity in MAP2-positive dendrites.

(D) The ratio of neuritic to somatic RTN4 intensity (N/S).

(E) The ratio of neuritic to somatic KDEL intensity (N/S).

All Data are plotted as mean±s.e.m. (B-D) Control: N = 3, n = 18; Rab10 KD: N = 3, n = 18; (B-D) A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, *** = p<0.001, ** = p<0.01, * = p <0.05.

Impaired ER dynamics in Rab10 KD neurons.

(A) Representative time-lapse of ER-mCherry3 signal before (upper), upon (middle) and after (bottom) photobleaching.

(B) Average normalized ER-mCherry3 fluorescence recovery after photobleaching in control and Rab10 KD hippocampal neurons.

(C) Normalized ER-mCherry3 fluorescence recovery after photobleaching at T = 220 s in control and Rab10 KD hippocampal neurons.

All Data are plotted as mean±s.e.m. (B, C) Control: N = 3, n = 23; Rab10 KD: N = 3, n = 23; (B, C) A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001.

Rab10 depletion does not increase ER stress.

(A) Representative images of wild-type (WT) neurons treated with vehicle (top) or tunicamycin (TM, middle) and Rab10 KD neurons treated with vehicle (bottom). Neurons were stained for ATF4 and MAP2. Scale bar: 50 μm.

(B) Quantification of ATF4 intensity in soma from each condition. All data are presented as mean±s.e.m. WT + vehicle: N = 2, n = 25; WT + vehicle: N = 2, n = 30; Rab10 KD + vehicle: N = 2, n = 14. A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, ns = not significant.

Tubular ER was also observed in some presynaptic sections, consistent with previous studies (Deng et al., 2021; Droz et al., 1975; Wu et al., 2017). The percentage of synaptic sections containing tubular ER was decreased by 24% (37% in Control versus 28% in Rab10 KD). Rough ER (rER) was also studied in the somata of Rab10 depleted and control neurons (Figure 4I). The diameter of rER tubes was reduced by 15% in Rab10 KD neurons. Hence, the substantial dysregulation of ribosomal and ER proteins in Rab10-depleted neurons was accompanied by changes in the abundance of synaptic ER and small changes in ER morphology in the soma.

To study these effects on ER abundance and morphology further, we performed immunofluorescence staining for two endogenous ER markers, KDEL and RTN4. The average fluorescence intensity of RTN4 and KDEL staining was significantly decreased by 35% and 25% respectively in Rab10 KD neurons (Figure 4—figure supplement 1A-C). The relative distributions of RTN4 and KDEL in neurites, as calculated by the intensity ratio of these two proteins in neurites over their somatic intensity, were reduced by 25% and 13%, respectively (Figure 4—figure supplement 1D-E). In conclusion, the ultrastructural changes in ER abundance and morphology upon Rab10 depletion were accompanied by altered distribution of axonal ER, without affecting the ultrastructure of SVs and DCVs.

Finally, given the dynamic nature of ER tubular networks and the involvement of Rab10 in ER tubule extension in COS-7 cells (English and Voeltz, 2013), we tested ER dynamics in Rab10 KD and control neurons expressing the luminal ER marker mCherry-ER3 using live-cell imaging at DIV8, and performed FRAP experiments at DIV14. The recovery of mCherry-ER3 intensity after photobleaching was significantly slower in Rab10 KD neurons with only 50% recovery within three minutes, compared to 80% recovery in control neurons (Figure 4—figure supplement 2C). Collectively, these data indicate that Rab10 depletion leads to reduced levels of ER-resident proteins altered ER abundance and morphology, and impaired ER dynamics.

Rab10 regulates SERCA2 levels and ER Ca2+ homeostasis

The ER is the largest internal Ca2+ source in neurons and plays a crucial role in maintaining neuronal Ca2+ homeostasis (Karagas and Venkatachalam, 2019). The maintenance of ER Ca2+ primarily relies on the Sarco Endoplasmic Reticulum Calcium ATPase (SERCA), a Ca2+ pump. Among the three major paralogs of SERCA, SERCA2 is particularly enriched in neurons (Britzolaki et al., 2018; Periasamy and Kalyanasundaram, 2007; Xu and Van Remmen, 2021). Consistent with the proteomic analysis which revealed a reduced SERCA2 expression upon Rab10 depletion (Figure 3), immunoblotting confirmed the reduction of SERCA2 levels, showing a 50% reduction in Rab10 KD neurons (Figure 5A-5B). Therefore, ER alternations in Rab10 KD neurons may disrupt Ca2+ homeostasis, which is essential for DCV exocytosis.

Reduced SERCA2 levels and impaired ER Ca2+ homeostasis in Rab10 KD neurons.

(A) Typical immunoblot showing reduced SERCA2 levels in Rab10 KD hippocampal neurons.

(B) Quantification of protein levels in Rab10 KD neurons normalized to control.

(C) Quantification of somatic ER Ca2+ concentration.

(D) Quantification of dendritic ER Ca2+ concentration.

(E) Representative image of a neuron infected with ER-GCaMP6-150 displayed with a pseudo line. Scale bar: 3 μm.

(F) Typical kymographs of the somatic intensity of ER-GCaMP6-150 showing the intensity decrease upon caffeine superfusion (red line), and the recovery in intensity after caffeine washout. Scale bar: 10 s.

(G) Average normalized ER-GCaMP6-150 fluorescence recovery after caffeine treatment.

(H) Normalized ER-GCaMP6-150 fluorescence recovery after caffeine treatment at T = 190 s. All Data are plotted as mean±s.e.m. (B) Control: N = 4, n = 4; Rab10 KD: N = 4, n = 4; (C-D) Control: N = 3, n = 17; Rab10 KD: N = 3; n = 17; Rescue: N = 3, n = 17 (H) Control: N = 3, n = 23; Rab10 KD: N = 3; n = 24; GDP-Rab10: n =3, n = 10; Rescue: N = 3, n = 24. A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, *** = p<0.001, ** = p<0.01, ns = not significant.

Caffeine triggers less ER Ca2+ release in Rab10 KD neurons.

(A) Left: representative cytosolic Fluo-5 AM signals upon Caffeine perfusion. Right: representative kymographs of cytosolic Fluo-5 AM signals upon caffeine perfusion in somas.

(B) Average traces of Fluo-5 AM signals.

(C) Quantification of the peak values of the Fluo-5 AM fluorescence traces upon caffeine perfusion.

(D) Quantification of the area under the curve (AUC) of the Fluo-5 AM fluorescence traces upon caffeine perfusion.

All Data are plotted as mean±s.e.m. (C, D) Control: N = 3, n = 44; Rab10 KD: N = 3, n = 35; (C, D) A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, ** = p<0.01.

To test this, we next measured the Ca2+ concentration in ER ([Ca2+]ER) using the ER Ca2+ indicator ER-GCaMP6 (de Juan-Sanz et al., 2017). We observed a reduction in [Ca2+]ER in soma from 130 μM in control neurons to about 70 μM in Rab10-depleted neurons (Figure 5C). [Ca2+]ER in neuritis was also reduced by 15% in Rab10 KD neurons (Figure 5D). The reduction of [Ca2+]ER was rescued by the expression of a knockdown-resistant Rab10 construct. To validate this observation, we assessed ER Ca2+ homeostasis indirectly by measuring the effect of caffeine on cytosolic Ca2+ concentration. As expected, in the absence of extracellular Ca2+, caffeine application (1 μM) triggered an increase in cytosolic Ca2+ due to Ca2+ release from the ER in both wild-type and Rab10 depleted neurons. However, the peak and the area of the caffeine-induced Ca2+ response curves were both reduced by 30% in Rab10 KD neurons (Figure 5—figure supplement 1).

Furthermore, we examined ER Ca2+ dynamics following a 10-minute caffeine treatment. Caffeine activates the ryanodine receptor (RyR), leading to the depletion of ER Ca2+ (Endo, 1975; Fujimoto et al., 1980). As expected, perfusion with caffeine-induced ER Ca2+ depletion, followed by recovery towards pre-stimulation levels (Figure 5F-G). In wild-type neurons, Ca2+ levels were recovered by 90% at T=190 s. However, the refilling of ER Ca2+ was significantly delayed in Rab10 KD neurons or GDP-Rab10 expressing neurons. Ca2+ levels were only recovered by 50% at T=190 s (Figure 5G-H) in these neurons.

Finally, to investigate the consequence of the ER Ca2+ depletion on neuronal Ca2+ homeostasis in Rab10 KD neurons, we measured cytosolic Ca2+ responses triggered by action potentials (AP) using the Ca2+ indicator Fluo5F (Figure 6A-C) and the genetically encoded Synaptophysin-GCaMP6 (Figure 6D-F). The AP-induced Ca2+ responses in the soma, as measured by Fluo5F, were reduced by 40% upon Rab10 depletion (Figure 6C). Similarly, the AP-induced Ca2+ responses in presynaptic nerve terminals, measured by Synaptophysin-GCaMP6, were also reduced (20%, Figure 6F), although this effect was smaller than the effects of Rab10 KD on ER Ca2+ levels and Caffeine-induced ER Ca2+ depletion (Figure 5). Taken together, these data suggest that Rab10 knockdown leads to ER Ca2+ depletion and impairs neuronal Ca2+ homeostasis, which may be attributed to the reduced levels of SERCA2 level and slower ER Ca2+ refilling and may contribute to the impaired DCV exocytosis.

Impaired neuronal Ca2+ influx triggered by electrical stimulation.

(A) Representative time-lapse of cytosolic Fluo-5 AM upon electrical stimulation (16 APs, 50Hz) in somas of hippocampal neurons. Scale bar: 10 μm.

(B) Average normalized response of somatic Fluo-5 AM fluorescence upon stimulation (16 APs, 50Hz) in hippocampal neurons.

(C) Quantification of the area under the curve (AUC) of the Fluo-5 AM fluorescence traces.

(D) Typical neurons infected with Synaptophysin-GCaMP6 (upper), typical kymograph of a neurite (bottom) showing Synaptophysin-GCaMP6 intensity increase upon electrical stimulation (16 APs, 50Hz, blue bars). Scale bar: 5μm.

(E) Average normalized response of Synaptophysin-GCaMP6 fluorescence intensity at presynaptic boutons upon stimulation (16 APs, 50Hz) in hippocampal neurons.

(F) Quantification of the area under the curve (AUC) of the Synaptophysin-GCaMP6 fluorescence traces in control and Rab10 KD neurons.

All Data are plotted as mean±s.e.m. (C) Control: N = 4, n = 24; Rab10 KD: N = 4, n = 30; Rescue: N = 4, n = 27. (F) Control: N = 3, n = 33; Rab10 KD: N = 3; n = 27. A one-way ANOVA tested the significance of adding experimental group as a predictor. ** = p<0.01, * = p <0.05, ns = not significant.

Rab10 depletion impairs ionomycin-induced DCV exocytosis

To determine whether the impaired Ca2+ signaling explains the DCV exocytosis deficiency in Rab10 KD neurons, we stimulated DCV exocytosis using the Ca2+ ionophore ionomycin. This approach bypasses cellular Ca2+ homeostasis and artificially increases the intracellular Ca2+ concentration enough to trigger DCV exocytosis (Persoon et al., 2019). Unexpectedly, DCV exocytosis was still reduced in Rab10 KD neurons (Figure 7A-B). Although the reduction was substantial, 45%, this impairment was not as substantial as observed for action potential-induced DCV exocytosis (65%, Figure 2K). The total number of DCV was nearly identical between control and Rab10 KD neurons (Figure 7D). Thus, although a minor fraction of the DCV exocytosis deficits may be explained by impaired Ca2+ signaling (difference between 45% and 65%), other deficits explain most of the DCV exocytosis deficiency in Rab10 KD neurons.

Impaired DCV fusion induced by ionomycin in Rab10 KD neurons.

(A) Representative neurons during electrical stimulation superimposed with NPY-pHluorin fusion events (green dots). Scale bar: 10 μm.

(B) Cumulative plot of DCV fusion events per cell.

(C) Fraction of NPY-pHluorin-labeled DCVs fusing during stimulation.

(D) The total number of DCVs (total pool) of neurons analyzed in Figure 7B, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

All Data are plotted as mean±s.e.m. (C, D) Control: N = 3, n = 20; Rab10 KD: N = 3, n = 21. (C, D) A one-way ANOVA tested the significance of adding experimental group as a predictor. * = p <0.05, ns = not significant.

Rab10 regulates neuronal protein synthesis

Since significant dysregulation of ER markers and ribosomal proteins was observed upon Rab10 depletion, we investigated the effects of Rab10 on protein synthesis using SUnSET to detect nascent peptides formed during puromycin pulse labeling (Schmidt et al., 2009). SUnSET analysis revealed that global protein synthesis was reduced by 30% upon Rab10 depletion (Figure 8A-B).

Leucine supplementation ameliorates the deficits in protein synthesis and neuropeptide release in Rab10 KD neurons.

(A) Representative western blot showing puromycinilated proteins as a measure for de novo protein synthesis in each condition.

(B) Quantification of puromycin intensity in each condition.

(C) Representation of the DCV fusion assay. Leucine (5 μM) was added to the culture media and incubated for 72 h before DCV fusion assay. DMSO (1‰) was used as a control.

(D) Cumulative plot of DCV fusion events per cell.

(E) Fraction of NPY-pHluorin-labeled DCVs fusing during stimulation.

(F) The total number of DCVs (total pool) of neurons analyzed in Figure 8D, E, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

All Data are plotted as mean±s.e.m. (B) All: N = 3, n = 3 (E, F) Control: N = 3, n = 47; Control + leu: N = 3, n = 45; Rab10 KD: N = 3; n = 61; Rab10 + leu: N = 3, n = 54. Rab10 KD + Rab10: N = 3, n = (B) One sample t-test. (E, F) A one-way ANOVA tested the significance of adding experimental group as a predictor. ** = p<0.01, * = p <0.05, ns = not significant.

Rab10 depletion does not upregulate mTORC1 pathway

(A) Typical immunoblot showing pS6K1 levels in each condition.

(B) Quantification of relative pS6K1 levels in each condition. All Data are plotted as mean±s.e.m. (B) Control, Control + Leu: N = 2, n = 2, Rab10 KD, Rab10 KD + Leu: N = 2, n = 4.

Leucine supplementation does not rescue ER morphological deficiency in Rab10 KD neurons

(A) Typical examples showing the KDEL signals in each condition

(B) Quantification of RTN4 intensity in MAP2-positive dendrites.

(C) The ratio of neuritic to somatic RTN4 intensity (N/S).

All Data are plotted as mean±s.e.m. (B, C) Control: N = 3, n = 10; Rab10 KD: N = 3, n = 11; Rab10 KD + Leu: N = 3; n = 11. A one-way ANOVA tested the significance of adding experimental group as a predictor. **** = p<0.0001, ns = not significant.

Overexpression of SERCA2 does not resuce the DCV fusion deficits in Rab10 KD neurons.

(A) Typical examples showing the SERCA2 signals in each condition

(B) Cumulative plot of DCV fusion events per cell

(C) Summary graph of DCV fusion events per cell.

(D) Total number of DCVs (total pool) of neurons, measured as the number of NPY-pHluorin puncta upon NH4Cl perfusion.

(E) Fraction of NPY-pHluorin-labeled DCV fusing during stimulation.

All Data are plotted as mean±s.e.m. (C-E) Control: N = 2, n = 10; Rab10 KD: N = 2, n = 13; SERCA2 OE: N = 2; n = 15. A one-way ANOVA tested the significance of adding experimental group as a predictor. *** = p<0.001, ** = p<0.01, ns = not significant.

Protein synthesis impairments may be rescued by supplementation with leucine, a branched-chain amino acid, that promotes protein synthesis by activating the mTOR pathway (Ananieva et al., 2016). To test this in Rab10 KD neurons, additional L-leucine was added to the culture medium to increase the concentration to 5 mM for three days. Indeed, 5mM leucine supplementation significantly restored global protein synthesis deficits caused by Rab10 depletion (Figure 8A-B).

Finally, a similar impairment in global protein synthesis was observed when a loss-of-function mutant of Rab10 (Rab10T23N) was overexpressed in wild-type neurons (Figure 8A-B). The deficit in protein translation is unlikely attributable to the upregulated mTORC1 signaling as the relative phosphorylation level of pS6K1 was unaffected in Rab10 KD neurons (figure 8-figure supplement 1). Thus, Rab10 depletion or Rab10T23N expression reduces global protein synthesis in neurons, probably by dysregulation of ER and ribosomal function.

Leucine supplementation restores normal DCV exocytosis

We hypothesized that protein synthesis deficits in Rab10-depleted neurons explain most of the impaired DCV exocytosis (in addition to a minor fraction explained by disturbed Ca2+-homeostasis, see above) and tested whether leucine supplementation could rescue the DCV exocytosis deficits in Rab10 KD neurons. Rab10 depleted neurons expressing NPY-pHluorin were treated with 5mM leucine three days before live-cell imaging, or with dimethyl sulfoxide (DMSO) as a control. Leucine supplementation restored DCV exocytosis by 80% caused by Rab10 depletion but did not alter DCV exocytosis in control neurons (Figure 8C-F). However, leucine supplementation failed to rescue the defects in ER morphology in Rab10 KD neurons (figure 8-figure supplement 2).

These results suggest that impaired protein synthesis is a major factor contributing to DCV exocytosis deficits in Rab10-depleted neurons.

Discussion

In this study, we investigated the function of Rab10 in neuropeptide release in mature mouse hippocampal neurons. We found that DCV exocytosis triggered by action potential trains was reduced by 65% upon Rab10 depletion, whereas SV exocytosis was unaffected. In addition, we observed a depleted ER Ca2+ pool and an impaired action potential-induced Ca2+ response in Rab10 KD neurons. However, DCV exocytosis triggered by Ca2+-ionophore ionomycin, a triggering method independent of Ca2+ channels and internal Ca2+-stores, was also impaired, albeit to a lesser extent. Furthermore, ribosomal proteins were massively dysregulated, and protein synthesis was impeded upon Rab10 depletion. Finally, the DCV exocytosis deficit in Rab10 KD neurons was largely rescued by leucine supplementation. We conclude that the strong inhibition of DCV exocytosis upon Rab10 depletion is mostly due to protein synthesis deficiency and to a lesser extent by dysregulation of Ca2+ channels or internal Ca2+-stores.

Rab10 regulates neurite outgrowth but not membrane homeostasis in mature neurons

Rab10 is highly enriched in neurons and plays crucial roles in neuronal development (Taylor et al., 2015; Wang et al., 2011; Xu et al., 2014). Consistent with these previous findings, we observed a reduction of axonal and dendritic outgrowth in Rab10-depleted neurons. Also consistent with previous findings in invertebrates (Sasidharan et al., 2012), we observed that the endogenous levels of SV markers are unchanged, indicating that SV biogenesis was unaffected. However, morphological characterization of neurons infected with shRNAs after the first week in culture did not identify changes in the total length of dendrites or axons, indicating that membrane homeostasis in mature neurons was unaffected by Rab10 knockdown. Therefore, the strong deficit in neuropeptide release in DIV7-infected neurons is unlikely to be confounded by Rab10-dependent aspects of neuronal development.

Rab10 is crucial for DCV exocytosis

In C. elegans, neuropeptide release was abolished in Rab10 deletion mutants (Sasidharan et al., 2012). We observed a 60% reduction in neuropeptide release in Rab10-depleted mature mouse hippocampal neurons. The difference in effect size could be explained by the incomplete Rab10 depletion with shRNA silencing or by redundant pathways in mammals, e.g. Rab10 and Rab8 are closely related paralogs and share many common effectors (Homma et al., 2020).

Unlike Rab3, which travels together DCVs and exhibits a reduction of over 90% in DCV exocytosis in its quadruple knock-out neurons (Persoon et al., 2019), Rab10 does not typically travel with DCVs (Figure 2—figure supplement 3). In addition, no changes in DCV size, puncta intensity, puncta distribution, travel velocity, or distance were detected in Rab10 KD neurons. We conclude that DCV exocytosis deficiency in Rab10-depleted neurons is not caused by alterations in DCV biogenesis or transport and that Rab10 is not required for DCV trafficking.

Although Rab10 is found in subcellular fractions enriched in synaptic vesicles (Takamori et al., 2006; Taoufiq et al., 2020), it is dispensable for SV exocytosis. Evoked postsynaptic currents (EPSCs) were unaffected in Rab10 mutants in C. elegans (Sasidharan et al., 2012). In line with this, SV exocytosis was unaffected in Rab10-depleted hippocampal neurons. These results indicate that Rab10 is selectively required for DCV exocytosis, not for SV exocytosis. Strikingly, many synaptic proteins, including many involved in SV exocytosis, are among the most severely dysregulated proteins upon Rab10 depletion. SV exocytosis may be more resilient to acute protein changes (Sinha et al., 2011) than DCVs. In addition, vesicle secretion properties are different for DCVs and SVs. Unlike SVs, which are secreted upon a single electrical stimulation, DCVs need prolonged or more intense stimulation for the induction of fusion. Thus, the regulatory effects of Rab10 on DCV exocytosis might be amplified under prolonged stimulation.

Ca2+ homeostasis deficits contribute to DCV exocytosis deficits in Rab10 KD neurons

Rab10 KD neurons showed depleted ER Ca2+ and impaired cytosolic Ca2+ responses. These effects may contribute to the observed DCV exocytosis deficits. Ca2+ released from the ER promotes DCV mobility and potentiates neuropeptide release via activating the CaMKII pathway in Drosophila (Shakiryanova et al., 2007). However, this might not be the case in mouse neurons since most axonal ER takes Ca2+ up from cytosol, instead of releasing it (de Juan-Sanz et al., 2017). Secondly, triggering Ca2+ release from the ER did not alter DCV transport and fusion (unpublished data from our lab). Finally, a previous study from our lab has shown that CaMKII deficiency does not alter DCV exocytosis (Moro et al., 2020). Hence, dysregulation of ER Ca2+ dynamics may not directly explain the observed DCV exocytosis deficits in Rab10-depleted neurons. However, ER Ca2+ also regulates Ca2+ influx by modulating L-type voltage-gated Ca2+ channels at the plasma membrane via a STIM1-based feedback loop (de Juan-Sanz et al., 2017). Given the importance of Ca2+ influx for DCV exocytosis, dysregulation of ER Ca2+ dynamics may indirectly explain DCV exocytosis deficiency upon Rab10 depletion. However, Ionomycin-triggered DCV exocytosis, which bypasses voltage-gated Ca2+ channels, was still reduced, albeit to a lesser extent compared to action potential trains. This difference in effect size is consistent with a limited contribution for dysregulation of Ca2+ dynamics and voltage-gated Ca2+ channels to explain the impaired DCV exocytosis in Rab10 KD neurons, while the majority of this phenotype is explained by deficits downstream of protein synthesis.

A role of Rab10 in protein synthesis largely explains DCV exocytosis deficiency in Rab10 KD neurons

In line with the previous study (Lv et al., 2015), which indicated alteration in ER morphology in Rab10 KO embryonic cells. We observed impaired ER morphology and upregulated ribosomal proteins upon Rab10 depletion. The upregulation of ribosomal proteins might be a compensatory response to altered ER structure as mutation of other ER-shaping proteins, such as VCP and ALT1, causes similar ribosomal abnormalities as Rab10 depletion (Shih and Hsueh, 2016). Protein homeostasis is vital for neuronal function and deficiency in protein translation is related to several CNS disorders (Cajigas et al., 2010; Holt et al., 2019; Koga et al., 2011; Laguesse and Ron, 2020).

Dysregulation of ribosomes and ER is probably sufficient to explain the impaired protein translation in Rab10 KD neurons. Depleted ER Ca2+ may induce ER stress (Arruda and Hotamisligil, 2015; Fu et al., 2011) which may also contribute to protein translation inhibition, but ATF4 levels are unaffected in Rab10 KD neurons, suggesting that ER stress is at best limited (Figure 4—figure supplement 3A-B). In addition, Rab10 regulates the retrograde transport of TrkB signaling endosomes (Lazo and Schiavo, 2023), which may activate the CREB/mTOR pathway and promote protein synthesis (Moya-Alvarado et al., 2023). Hence, impaired TrkB transport may also contribute to impaired protein translation in addition to dysregulated ribosomes in Rab10 KD neurons. We found that restoring protein synthesis with leucine efficiently increased protein synthesis and largely restored DCV exocytosis deficiency in Rab10-depleted neurons. We conclude that dysregulation of protein synthesis results in DCV exocytosis deficits in these neurons.

In conclusion, our data demonstrate the importance of Rab10 in neuropeptide release and ER homeostasis. We observed altered ER morphology, reduced ER Ca2+ concentration, impaired protein synthesis, and impaired neuropeptide release in Rab10-depleted neurons. These observations shed light on the pathogenesis of Rab10-related disease. In addition, we have shown leucine can largely rescue deficiency in protein synthesis and neuropeptide release in Rab10-depleted neurons, providing a potential treatment for disorders associated with neuropeptide abnormalities such as depression and anxiety.

Contributions

Conceptualization: JD, RT, MV

Methodology: JD, MC, RD, JW, KWL, ABS, RT, MV

Investigation: JD, RT, MV

Visualization: JD, MC, KWL, JW, RT, MV

Funding acquisition: RT, MV

Supervision: RT, MV

Writing – original draft: JD, RT, MV